### 2012-03-07 abstract.tex
Begin work on abstract.

Nell is targeted at children far from educational infrastructure.

We present Project Nell, tablet software to teach reading and writing to
children far from educational infrastructure.  Nell is a modular
narrative system with ``a low floor and no ceilings'': the design
scales from teaching letter shapes to programming the system itself.

Nell is the successor to the Sugar educational software developed by
One Laptop per Child for their XO-1 laptop.  Nell targets the XO-3
tablet, a sub-\$100 solar-powered tablet for the developing world.
Experience with Sugar drove design choices for Nell.

\footnote Nell was inspired by the ``Young Ladies' Interactive
Primer'' in the Neal Stephenson novel, ``The Diamond Age''.  It takes
its name from novel's protagonist, but could also be read as an
acronym for ``Narrative Environment for Learning Learning'' (a nod to
Seymour Papert).

### 2012-03-07 paper.tex
Basic paper skeleton in ACM style.

% THIS IS SIGPROC-SP.TEX - VERSION 3.1
% WORKS WITH V3.2SP OF ACM_PROC_ARTICLE-SP.CLS
%
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission,
% you need to 'insert'  your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
\documentclass{acm_proc_article-sp}

\begin{document}

\title{Growing up with Nell: Narrative Interfaces for Literacy}
%\subtitle{Narrative Interfaces for Literacy}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{1}
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
C. Scott Ananian\\
\affaddr{One Laptop Per Child Foundation}\\
\affaddr{222 Third Street}\\
\affaddr{Cambridge, MA 02142}\\
\email{cscott@laptop.org}
%\and  % use '\and' if you need 'another row' of author names
}
\date{12 March 2012}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle

\begin{abstract}
We present Project Nell,%
\footnote{We were inspired by the ``Young Lady's Illustrated
Primer'' in the Neal Stephenson novel \textit{The Diamond Age}.
Nell takes its name from that novel's protagonist, but could also be read as an
acronym for ``Narrative Environment for Learning Learning'' (a nod to
Seymour Papert).}
tablet software to teach reading and writing to
children far from educational infrastructure.  Nell is a modular
narrative system with ``a low floor and no ceilings'': the design
scales from teaching letter shapes to programming the system itself.

Nell is the successor to the Sugar educational software developed by
One Laptop per Child for their XO-1 laptop.  Nell targets the XO-3
tablet, a sub-\$100 solar-powered tablet for the developing world.
Experience with Sugar drove design choices for Nell.
\end{abstract}

% A category with the (minimum) three required fields
\category{K.3.1}{Computers and Education}{Computer
  Uses in Education}[computer-managed instruction]
\category{I.2.7}{Artificial Intelligence}{Natural Language
  Processing}[narrative interfaces]
\category{H.5.2}{Information Interfaces and Presentation}{User
  Interfaces}[narrative interfaces]
\terms{Design, Human Factors}

\keywords{Narrative interfaces, tablet computing, education, Nell}

\section{Introduction}
\section{The {\secit Body} of The Paper}
\subsection{Type Changes and {\subsecit Special} Characters}
\subsection{Math Equations}
\subsection{Citations}
Citations to articles \cite{bowman:reasoning, clark:pct, braams:babel, herlihy:methodology},
conference
proceedings \cite{clark:pct} or books \cite{salas:calculus, Lamport:LaTeX} listed
in the Bibliography section of your
article will occur throughout the text of your article.

This article shows only the plainest form
of the citation command, using \texttt{{\char'134}cite}.
This is what is stipulated in the SIGS style specifications.
No other citation format is endorsed.

\section{Conclusions}
This paragraph will end the body of this sample document.
Remember that you might still have Acknowledgments or
Appendices; brief samples of these
follow.  There is still the Bibliography to deal with; and
we will make a disclaimer about that here: with the exception
of the reference to the \LaTeX\ book, the citations in
this paper are to articles which have nothing to
do with the present subject and are used as
examples only.
%\end{document}  % This is where a 'short' article might terminate

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
\balancecolumns
% That's all folks!
\end{document}

### 2012-03-07 paper.tex
Initial outline.

@@
 
-\title{Growing up with Nell: Narrative Interfaces for Literacy}
+\title{Growing up with Nell: A Narrative Interface for Literacy}
 %\subtitle{Narrative Interfaces for Literacy}
@@
 \begin{abstract}
-We present Project Nell,%
+Project Nell%
 \footnote{We were inspired by the ``Young Lady's Illustrated
@@
 Seymour Papert).}
-tablet software to teach reading and writing to
+is tablet software to teach reading and writing to
 children far from educational infrastructure.  Nell is a modular
@@
 
-\section{Introduction}
-\section{The {\secit Body} of The Paper}
-\subsection{Type Changes and {\subsecit Special} Characters}
-\subsection{Math Equations}
-\subsection{Citations}
-Citations to articles \cite{bowman:reasoning, clark:pct, braams:babel, herlihy:methodology},
-conference
-proceedings \cite{clark:pct} or books \cite{salas:calculus, Lamport:LaTeX} listed
-in the Bibliography section of your
-article will occur throughout the text of your article.
+\section{A Lesson from Nell}
+% tell a nell story up front
 
-This article shows only the plainest form
-of the citation command, using \texttt{{\char'134}cite}.
-This is what is stipulated in the SIGS style specifications.
-No other citation format is endorsed.
+\section{Key Ideas}
+% olpc context here?
+\cite{chang:tinkrbook, flores:uruguay, idb:peru}
+%context: english literacy, far from teachers & schools
+\subsection{Narrative}
+% excerpt of story model
+\subsection{Personal}
+% customization, renaming characters
+\subsection{Direct interaction}
+% handwriting as UI?
+% motor skills
+% photo of XO-3?
+\subsection{Growable (``no ceiling'')}
+% modular -- the story expands.
+% excerpt of TurtleScript
+% text read aloud initially
+
+\section{Technology}
+\subsection{Rule-based story modules}
+% rete algorithm, cite Riedl, etc.
+% programmable conflict resolution
+\subsection{Simple dialog with AIML}
+% eliza interactions, show fragment
+\subsection{Web technologies}
+% Javascript/HTML5
+% offline caching
+% turtles (almost) all the way down (turtlescript again)
+
+\section{Related work}
+\subsection{The Sugar Learning Environment}
+% Sugar
+%% customization
+%% path through the educational material?
+\subsection{Narrative computing}
+% Reidl
+\subsection{Storytelling}
+% TinkRBooks
 
 \section{Conclusions}
-This paragraph will end the body of this sample document.
-Remember that you might still have Acknowledgments or
-Appendices; brief samples of these
-follow.  There is still the Bibliography to deal with; and
-we will make a disclaimer about that here: with the exception
-of the reference to the \LaTeX\ book, the citations in
-this paper are to articles which have nothing to
-do with the present subject and are used as
-examples only.
-%\end{document}  % This is where a 'short' article might terminate
 
@@
 \bibliographystyle{abbrv}
-\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
+\bibliography{paper}  % paper.bib is the name of the Bibliography in this case
 % You must have a proper ".bib" file

### 2012-03-08 paper.tex
Intro story.

@@
 Project Nell%
-\footnote{We were inspired by the ``Young Lady's Illustrated
-Primer'' in the Neal Stephenson novel \textit{The Diamond Age}.
-Nell takes its name from that novel's protagonist, but could also be read as an
-acronym for ``Narrative Environment for Learning Learning'' (a nod to
-Seymour Papert).}
+
 is tablet software to teach reading and writing to
@@
 \section{A Lesson from Nell}
+
 % tell a nell story up front
+Miles from the nearest school, a young girl in Ethiopia turns on a
+tablet computer for the first time.  The solar-powered machine begins
+talking to her immediately: \textit{``Hello!  Would you like to hear a story?''}
+
+She listens intently to a story about a princess named Nell.  Later, when
+the girl has learned a little more, she will give the princess her own
+name, ``Rahel,'' and change all the colors, but for now the green
+book shows her pictures and asks her to trace shapes on the screen.
+
+\textit{``R is for Raven.  Can you trace the R?''}
+
+As she traces the R, it comes to life and gallops across the screen.
+\textit{``Run starts with R.  Roger the R Runs across the Red Rug.  He has a
+dog named Rover.''}  Rover barks: \textit{``Ruff! Ruff!''}  The Princess asks,
+\textit{``Can you find something Red?''} and Rahel uses the camera to
+photograph her red boots. \textit{``Very good!  You are a smart girl!''}
+
+As Rahel grows, the book asks her to trace not just letters, but whole
+words.   The books responses are written on the screen as it writes
+them, and eventually she doesn't need to leave the sound on all the
+time to read what the Princess is saying.  Soon Rahel can write
+complete sentences in her special book, and the Princess will respond
+sometimes.  New stories teach her about music (by playing certain
+tunes she unlocks a dungeon door) and programming with blocks (a
+not-very-bright turtle needs her to tell him how to draw different
+shapes).  Rahel writes her own stories, which she shares with her
+far-away friends.
+
+An older Rahel learns that the block language she used to talk with
+the turtle is actually used to write all the software running inside
+her special book.  The stories Rahel writes can now include new
+functionality that make her book even more exciting for the kids
+just now opening their green meta books.
 
 \section{Key Ideas}
+The interaction design of the Nell system described above is inspired
+by the ``Young Lady's Illustrated Primer'' in the Neal Stephenson
+novel \textit{The Diamond Age}, from whose protagonist Nell takes its
+name.\footnote{In a nod to Seymour Papert, Nell can also be read as an
+acronym for ``Narrative Environment for Learning Learning.''}
+Nell's design embodies four key ideas: it is a \textbf{Narrative}
+interface which is \textbf{Personalized} for the child and uses
+\textbf{Direct Interaction}.  Nell is \textbf{Growable}: it has a
+``low floor and no ceiling'' to allow a child to master more
+and more of the system as she grows.
+
 % olpc context here?

### 2012-03-08 paper.tex
Lead in to our four key ideas.

@@
 name, ``Rahel,'' and change all the colors, but for now the green
-book shows her pictures and asks her to trace shapes on the screen.
+book draws pictures for her and asks her to trace shapes on the screen.
 
@@
 \textit{``Can you find something Red?''} and Rahel uses the camera to
-photograph her red boots. \textit{``Very good!  You are a smart girl!''}
+photograph her red boots.
+
+\textit{``Very good!  You are a smart girl!''}
 
@@
 Nell's design embodies four key ideas: it is a \textbf{Narrative}
-interface which is \textbf{Personalized} for the child and uses
-\textbf{Direct Interaction}.  Nell is \textbf{Growable}: it has a
-``low floor and no ceiling'' to allow a child to master more
-and more of the system as she grows.
+interface using \textbf{Direct Interaction} which \textbf{Grows} with,
+and is \textbf{Personalized} for, the child.
 
-% olpc context here?
-\cite{chang:tinkrbook, flores:uruguay, idb:peru}
-%context: english literacy, far from teachers & schools
+We believe the combination of these four key ideas provides a novel
+experience which addresses some of the shortcomings of earlier
+learning environments for children.  In this section we will discuss
+our motivating core concepts, and in the next we will describe how
+they are implemented in Nell.
+
+%Section~\ref{sec:related} discusses related work.
+%The final section will draw conclusions and describe future work.
+
 \subsection{Narrative}
@@
 \subsection{Growable (``no ceiling'')}
+% it has a ``low floor and no ceiling'' to allow a child to master more
+% and more of the system as she grows.
+
 % modular -- the story expands.
@@
 
-\section{Related work}
+\section{Related work}\label{sec:related}
+% Nell focused on Literacy
+% olpc context here?
+\cite{chang:tinkrbook, flores:uruguay, idb:peru}
+%context: english literacy, far from teachers & schools
 \subsection{The Sugar Learning Environment}

### 2012-03-08 paper.tex
Start writing "Narrative Interface" section.

@@
 \begin{abstract}
-Project Nell%
-
-is tablet software to teach reading and writing to
+Project Nell is tablet software to teach reading and writing to
 children far from educational infrastructure.  Nell is a modular
@@
 tablet, a sub-\$100 solar-powered tablet for the developing world.
-Experience with Sugar drove design choices for Nell.
+%Experience with Sugar drove design choices for Nell.
+
+\textbf{XXX REWRITE THE ABSTRACT XXX}
 \end{abstract}
@@
 As Rahel grows, the book asks her to trace not just letters, but whole
-words.   The books responses are written on the screen as it writes
+words.  The book's responses are written on the screen as it speaks
 them, and eventually she doesn't need to leave the sound on all the
-time to read what the Princess is saying.  Soon Rahel can write
-complete sentences in her special book, and the Princess will respond
-sometimes.  New stories teach her about music (by playing certain
-tunes she unlocks a dungeon door) and programming with blocks (a
-not-very-bright turtle needs her to tell him how to draw different
-shapes).  Rahel writes her own stories, which she shares with her
-far-away friends.
+time.  Soon Rahel can write complete sentences in her special book,
+and the Princess will respond sometimes.  New stories teach her about
+music (by playing certain tunes she unlocks a dungeon door) and
+programming with blocks (a not-very-bright turtle needs Princess Rahel's
+help to draw different shapes).  Rahel writes her own stories,
+which she shares with her friends.
 
 An older Rahel learns that the block language she used to talk with
-the turtle is actually used to write all the software running inside
+the turtle is also used to write all the software running inside
 her special book.  The stories Rahel writes can now include new
@@
 learning environments for children.  In this section we will discuss
-our motivating core concepts, and in the next we will describe how
+our motivating concepts, and in the next we will describe how
 they are implemented in Nell.
@@
 
-\subsection{Narrative}
+\subsection{Narrative Interface}
+
+Experience with our previous educational software
+system~\cite{flores:uruguay,idb:peru} has shown gratifying uptake and
+use by children.  But we are often disappointed that some of our
+favorite pedagogical tools included in the systems were not being
+widely used.  Enthusiastic teachers would spark intensive use in
+certain classrooms, but children had difficulty finding their way into
+the material without strong guidance.
+
+Children (like all humans) are hard-wired for
+stories~\cite{boyd:stories}.  Nell uses a
+\textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative,laurel:computers} 
+to guide the child user through the pedagogical capabilities of the system.
+
 % excerpt of story model
+% not always a story -- kid can draw on their own and return to the
+% story; Nell will observe the free play and offer suggestions.
 \subsection{Personal}
@@
 % olpc context here?
-\cite{chang:tinkrbook, flores:uruguay, idb:peru}
+\cite{chang:tinkrbook}
 %context: english literacy, far from teachers & schools
@@
 \section{Conclusions}
+
+\section{Acknowledgments}
+Nell's design evolved from experience with the Sugar system,
+architected by Walter Bender, and through stimulating discussions with
+Chris Ball and Angela Chang.  Michael Stone also contributed to an
+early implementation.
 

### 2012-03-08 paper.tex
Finish 1st draft of 'Narrative Interface' section under 'Key Ideas.

@@
 \textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative,laurel:computers} 
-to guide the child user through the pedagogical capabilities of the system.
+to guide the child user through the pedagogical capabilities of the
+system.  All system actions are shaped by the storybook metaphor, and
+interactions with the system are conceptualized as interactions with
+one of the system protagonists.
 
-% excerpt of story model
+The overall system story is a multi-character serial adventure, taking
+cues from \textit{The Diamond Age} and
+from the UNIVERSE narrative-generation system \cite{lebowitz:universe}.
+Each character represents a specific skill or subject area, and each
+adventure represents about a year's curriculum.  This design also
+improves modularity and allows updates and decentralized authorship.
+
+As shown in figure~\ref{fig:nell}, Nell's characters are
+always-available agents running on top of a system ``activity''
+(application) which provides specialized functionality.  The agents
+are not always foregrounded; the child is always free to engage in
+constructionist play within the base activity.  The handwriting tutor is
+fundamentally a drawing activity; the adventure involving the magical
+musical lock is also a music-making activity.  The narrative system is
+hooked into the activity to provide passive guidance
+(congratulating the child when it notices they've drawn a letter),
+active guidance (Apple-Guide--style~\cite{powers:appleguide}
+contextual help), or system services (switch activities; jump into a
+related story module/lesson plan).
+
+% Story Module == Lesson Plan
+
 % not always a story -- kid can draw on their own and return to the
@@
 % customization, renaming characters
+
+Kids love to personalize stuff \textbf{[CITATION NEEDED]}.
+
 \subsection{Direct interaction}
@@
 \subsection{Rule-based story modules}
+% excerpt of story model
 % rete algorithm, cite Riedl, etc.

### 2012-03-08 paper.tex
Rough in 'Personalization' section.

@@
 % story; Nell will observe the free play and offer suggestions.
-\subsection{Personal}
+\subsection{Personalization}
 % customization, renaming characters
 
-Kids love to personalize stuff \textbf{[CITATION NEEDED]}.
+Children often decorate their belongings, and the
+child owning a Nell system is encouraged to personalize it.
+Superficially, colors can be changed and characters can be renamed and
+dressed up.  But Nell also supports story personalization at a deeper
+level.
+
+Choices the child has made are stored in an \textit{Avatar Services}
+module, accessible to all story modules and activities.  Past choices
+can thus influence the future direction of a story.  Story module
+alternates are supported to allow learning material to presented in
+different ways for different learners. Further, the record of a
+child's experiences with Nell are recorded in a \textit{Journal},
+which can be edited and customized by the user.  Careful arrangement
+of sections of the Journal can become a Portfolio \textbf{cite here}.
+The journal also allows undo/redo \textbf{XXXX}.
+
+In the future, we expect to allow further personalization via deep
+recombination of story elements.  Lebowitz \textbf{PLANNING} and Riedl
+\textbf{PLANNING} show how a planner can be used to recombine story
+modules.  We expect to be less ambitious then previous work: instead
+of attempting to generate thousands of stories from tens of templates,
+we expect to modestly adapt from hundreds of story modules created
+in a decentralized manner by teachers.
 
 \subsection{Direct interaction}
+% constructionist?
 % handwriting as UI?
@@
 % photo of XO-3?
+
 \subsection{Growable (``no ceiling'')}
@@
 \end{document}
+
+%%  LocalWords:  personalization

### 2012-03-08 paper.tex
Updates & edits.

--- 2012-03-08 paper.tex
@@
 % tell a nell story up front
-Miles from the nearest school, a young girl in Ethiopia turns on a
+Miles from the nearest school, a young Ethiopian girl turns on a
 tablet computer for the first time.  The solar-powered machine begins
@@
 
-She listens intently to a story about a princess named Nell.  Later, when
-the girl has learned a little more, she will give the princess her own
-name, ``Rahel,'' and change all the colors, but for now the green
-book draws pictures for her and asks her to trace shapes on the screen.
+She nods and listens to a story about a princess.  Later, when
+the girl has learned a little more, she will tell the machine that the
+princess is named ``Rahel,'' like she is, and change all its
+colors---but for now the green book draws pictures of the unnamed Princess
+for her and asks her to trace shapes on the screen.
 
@@
 As she traces the R, it comes to life and gallops across the screen.
-\textit{``Run starts with R.  Roger the R Runs across the Red Rug.  He has a
+\textit{``Run starts with R.  Roger the R runs across the Red Rug.  Roger has a
 dog named Rover.''}  Rover barks: \textit{``Ruff! Ruff!''}  The Princess asks,
 \textit{``Can you find something Red?''} and Rahel uses the camera to
-photograph her red boots.
+photograph her new boots.
 
@@
 time.  Soon Rahel can write complete sentences in her special book,
-and the Princess will respond sometimes.  New stories teach her about
-music (by playing certain tunes she unlocks a dungeon door) and
-programming with blocks (a not-very-bright turtle needs Princess Rahel's
-help to draw different shapes).  Rahel writes her own stories,
+and sometimes the Princess will respond to them.  New stories teach her about
+music (she unlocks a dungeon door by playing certain tunes) and
+programming with blocks (Princess Rahel helps a not-very-bright turtle
+to draw different shapes).  Rahel writes her own stories about the Princess,
 which she shares with her friends.
@@
 functionality that make her book even more exciting for the kids
-just now opening their green meta books.
+just now receiving their own green meta books.
 
@@
 
-We believe the combination of these four key ideas provides a novel
+We believe the combination of these four key concepts provides a novel
 experience which addresses some of the shortcomings of earlier
 learning environments for children.  In this section we will discuss
-our motivating concepts, and in the next we will describe how
-they are implemented in Nell.
+these four building blocks, and in the next section we will describe their
+implementation in Nell.
 
@@
 \textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative,laurel:computers} 
-to guide the child user through the pedagogical capabilities of the system.
+to guide the child user through the pedagogical capabilities of the
+system.  All system actions are shaped by the storybook metaphor, and
+interactions with the system are conceptualized as interactions with
+one of the system protagonists.
 
-% excerpt of story model
+Nell's overall story is a multi-character serial adventure, taking
+cues from \textit{The Diamond Age} and
+from the UNIVERSE narrative-generation system \cite{lebowitz:universe84}.
+Each of the several characters represents a specific skill or subject
+area, and each adventure represents about a year's curriculum.  This
+design also improves modularity and allows updates and decentralized
+authorship.
+
+As illustrated in figure~\ref{fig:nell}, Nell's characters are
+always-available agents running on top of a particular system \textit{activity}
+(application) which provides specialized functionality.  The agents
+are not always foregrounded: constructionist learning occurs when the
+child plays freely to ``make things'' with the base activity.
+The handwriting tutor is fundamentally a drawing activity; the
+adventure involving the magical musical lock is also a music-making activity.
+%  ``Making things'' is the key to constructionist learning.
+The narrative system is
+hooked into each activity to provide passive guidance
+(congratulating the child when it notices they've drawn a letter),
+active guidance (Apple-Guide--style~\cite{powers:appleguide}
+contextual help), or system services (switch activities; jump into a
+related story module/lesson plan).
+
+% Story Module == Lesson Plan
+
 % not always a story -- kid can draw on their own and return to the
 % story; Nell will observe the free play and offer suggestions.
-\subsection{Personal}
+\subsection{Personalization}
 % customization, renaming characters
+
+Children often decorate their belongings, and the
+child owning a Nell system is encouraged to personalize it.
+Superficially, colors can be changed and characters can be renamed and
+dressed up.  But Nell also supports deeper story personalization.
+
+Choices the child has made are stored in \textit{Avatar Services},
+accessible to all story modules and activities.  Past choices
+can thus influence the future direction of a story.  Story module
+alternatives are supported to allow learning material to presented in
+different ways for different learners. Further, the record of a
+child's experiences with Nell are recorded in a \textit{Journal},
+which can be edited and customized by the user.  Careful arrangement
+of sections of the Journal can become a Portfolio~\cite{stefanakis:portfolios}.
+The journal also allows undo/redo \textbf{XXXX}.
+
+In the future, we expect to allow further personalization via deep
+recombination of story elements.  Lebowitz~\cite{lebowitz:universe85} and Riedl~\cite{riedl:planning} show how a planner can be used to recombine story
+modules.  We expect to be less ambitious then previous work: instead
+of attempting to generate thousands of stories from tens of templates,
+we expect to modestly adapt from hundreds of story modules created
+in a decentralized manner by teachers.
+
 \subsection{Direct interaction}
+% constructionist?
 % handwriting as UI?

### 2012-03-09 paper.tex
Journal, Multiple Intelligences, Portfolios.  Finish the Personalization draft.

@@
 
-Choices the child has made are stored in \textit{Avatar Services},
-accessible to all story modules and activities.  Past choices
-can thus influence the future direction of a story.  Story module
-alternatives are supported to allow learning material to presented in
-different ways for different learners. Further, the record of a
-child's experiences with Nell are recorded in a \textit{Journal},
-which can be edited and customized by the user.  Careful arrangement
-of sections of the Journal can become a Portfolio~\cite{stefanakis:portfolios}.
-The journal also allows undo/redo \textbf{XXXX}.
+\textit{Avatar Services} stores all choices made by the child and make them
+accessible to story modules and activities.  Accomplishments
+(completed lessons) are also stored.  Nell can thus provide multiple
+alternative story modules for a given plot point/lesson, and select
+between them based on either the explicit choices made by the child
+(``fractions in outer space,'' if the child previously indicated a
+preference for space-themed lessons) or prior success with a given
+lesson style (a large number of accomplishments in musical-rhythmic
+tasks suggests a rhythmic approach to fractions).  In this manner we
+can engage the child's interests and cater to different learning
+styles~\cite{gardner:mi}.
 
-In the future, we expect to allow further personalization via deep
-recombination of story elements.  Lebowitz~\cite{lebowitz:universe85} and Riedl~\cite{riedl:planning} show how a planner can be used to recombine story
-modules.  We expect to be less ambitious then previous work: instead
-of attempting to generate thousands of stories from tens of templates,
-we expect to modestly adapt from hundreds of story modules created
-in a decentralized manner by teachers.
+Moving further, this record of past choices and accomplishments can be
+reviewed.  The narrative history is stored in the \textit{Journal},
+where it can form the core of a Portfolio~\cite{stefanakis:portfolios}.
+In order to encourage fearless play and experimentation, the Journal
+also supports pervasive undo; you can replay the narrative starting at
+any past point in the journal.
+
+In the future, we expect to allow further narrative personalization
+via recombination of story elements.
+Lebowitz~\cite{lebowitz:universe85} and Riedl~\cite{riedl:planning}
+show how a planner can be used to recombine and adapt story fragments.
+In this we expect to be less ambitious than the cited work: instead of
+attempting to generate thousands of stories from tens of templates, we
+expect to select and then modestly adapt from hundreds of story
+modules created in a decentralized manner by teachers---and eventually
+by the students themselves.
 
@@
 
-%%  LocalWords:  personalization
+%%  LocalWords:  personalization Lebowitz

### 2012-03-09 paper.tex
First draft of 'Growable' section.

@@
 
-\subsection{Narrative Interface}
+\subsection{Narrative}
 
@@
 % story; Nell will observe the free play and offer suggestions.
-\subsection{Personalization}
+\subsection{Personalized}
 % customization, renaming characters
@@
 
+\subsection{Growable}
+The Nell system aims to have a ``low floor and no ceiling.''  The low
+floor is a very simple base system, usable with no literacy skills or
+external guidance.  The system then grows with the child's expanding
+capabilities, serially accumulating story modules to tackle topics of
+increasing difficulty.  As mentioned before, the ability for third
+parties to independently author story modules is necessary to ensure
+Nell can continue to grow.  There should be no external dependencies
+limiting a child's ability to participate in this process: a Nell
+system should contain a story editor and everything necessary to
+author and publish story modules.
+
+We also seek to eliminate the ``ceiling'' between the children and
+the authors of the Nell system itself.  Nell should be capable of
+teaching about itself, and a capable child can dive into Nell's source
+code and make meaningful changes.  In section~\ref{sec:turtles} we
+describe how this goal is approached.
+
+% it has a ``low floor and no ceiling'' to allow a child to master more
+% and more of the system as she grows.
+% modular -- the story expands.
+% text read aloud initially
+
 \subsection{Direct interaction}
@@
 % photo of XO-3?
-
-\subsection{Growable (``no ceiling'')}
-% it has a ``low floor and no ceiling'' to allow a child to master more
-% and more of the system as she grows.
-
-% modular -- the story expands.
-% excerpt of TurtleScript
-% text read aloud initially
+The Nell system is tied to its implementation on a tablet computer.
+It's no accident that the Nell system is instantiated on a tablet.
+XXX
 
@@
 % eliza interactions, show fragment
-\subsection{Web technologies}
+\subsection{Web technologies}\label{sec:turtles}
 % Javascript/HTML5
@@
 % turtles (almost) all the way down (turtlescript again)
+% excerpt of TurtleScript
 

### 2012-03-09 paper.tex
False starts on 'direct interaction' section.

@@
 limiting a child's ability to participate in this process: a Nell
-system should contain a story editor and everything necessary to
+system contains a story editor and everything necessary to
 author and publish story modules.
@@
 % photo of XO-3?
+
 The Nell system is tied to its implementation on a tablet computer.
+
 It's no accident that the Nell system is instantiated on a tablet.
-XXX
+
+The constructionist learning philosophy seeks to engage kids by
+``making things''.  We believe that the direct interaction afforded by
+tablet computing helps approach that goal.
+
+Nell is deployed on a tablet computer, and...
+
+Nell's implementation on a tablet computer advances...
+
+\textbf{XXX}
 

### 2012-03-09 paper.tex
Write "Direct Interaction" section; stub out figures.

@@
 \section{A Lesson from Nell}
+\begin{figure}
+\centering
+\epsfig{file=roger1.eps, height=1.5in} % width=3.2in
+\caption{Illustration of Nell's primary interface.}\label{fig:nell}
+\end{figure}
 
@@
 As illustrated in figure~\ref{fig:nell}, Nell's characters are
-always-available agents running on top of a particular system \textit{activity}
+always-available agents layered above a particular system \textit{activity}
 (application) which provides specialized functionality.  The agents
@@
 
-The Nell system is tied to its implementation on a tablet computer.
+\begin{figure}
+\centering
+\epsfig{file=xo3-3.eps, width=3.2in}
+\caption{Nell's solar-powered hardware platform.  The tablet design
+  allows for direct interaction.}\label{fig:xo3}
+\end{figure}
 
-It's no accident that the Nell system is instantiated on a tablet.
+The constructionist learning philosophy emphasizes creation and
+tangible interaction.  Nell uses direct interaction on a tablet
+computer to maintain the child's connection to their work
+(Figure~\ref{fig:xo3}).
 
-The constructionist learning philosophy seeks to engage kids by
-``making things''.  We believe that the direct interaction afforded by
-tablet computing helps approach that goal.
-
-Nell is deployed on a tablet computer, and...
-
-Nell's implementation on a tablet computer advances...
-
-\textbf{XXX}
+The direct interaction model lowers the floor by eliminating the need
+to learn an abstract touchpad or mouse interface.  It also supports
+our literacy goals by allowing direct handwriting instruction and the
+use of handwriting as an interface.
 

### 2012-03-09 paper.tex
Rough out the Technology section.

@@
 \section{Technology}
+We have begun implementation of the Nell system.  In this section we
+will describe the implementation of three of the most novel pieces
+from the software engineering side.  These are not necessarily the
+most novel-appearing from the user's perspective: for example, the
+handwriting UI is still fairly unusual in consumer devices, but the
+implementation of a separated-character handwriting recognition engine
+using Hidden Markov Models is well-understood.
+\textbf{SMOOTH THIS LANGUAGE}
+
 \subsection{Rule-based story modules}
+Narrative is the core of the Nell interface.  It is important that
+Nell's story be modular and extensible.  We use a rule-based story
+model with programmable conflict resolution to provide this ability.
+Figure~\ref{fig:rules} shows an excerpt of the story module driving
+the interaction described in the introduction to this paper.
+
+XXX stories consist of elements.  all elements contain preconditions,
+and actions.  cite PLANNER.
+
+In principle, all rules for all elements of all stories are evaluated
+continuously to determine the actions of the narrative agents.  This
+is made computationally efficient using the Rete
+algorithm~\cite{rete}.  The Rete algorithm requires that our
+conditions are in Normal Form\footnote{BE PRECISE HERE}, but actions
+can be arbitrarily complex.
+
+New story modules can be incorporated by visiting an internet page in
+connected deployments.  In a disconnected deployments, new stories
+might be distributed once a year on USB sticks, or shared from your
+friends.  In a classroom environment, a teacher can share stories for
+the day's lesson at the start of class.
+
+Story modules can be inherited and extended, as figure~\ref{fig:rules2}
+shows.  This makes it easy to take an existing story and modify it to
+better suit particular interests or a particular type of learner, or
+just to add variety.  This can lead to conflicts: multiple versions of
+a story, or multiple variant elements within a story, may have their
+preconditions satisfied simultaneously.
+
+We resolve conflicts...
+
+We have not yet implemented an integrated story editor within Nell,
+but we expect it to be roughly similar to the one implemented for
+Wide Ruled, which showed good learnability by inexperienced
+users~\cite{wideruled}.
+
 % excerpt of story model
@@
 % programmable conflict resolution
-\subsection{Simple dialog with AIML}
+% cite the story editor from Wide Ruled
+
+\subsection{Extensible dialog with AIML}
+The child can interact with the Princess, and other agents within
+Nell, directly.  Convincing discourse is a hard problem, but
+reasonable approximations do not have to be difficult.  We leverage
+the AIML language developed for AliceBot~\cite{xxx}, which has
+won several Turing contests. \textbf{XXX more precise here XXX}
+\footnote{Anecdotal evidence shows that interactions with Eliza-style
+  responders can be rewarding: one woman spent hours each day
+  conversing with the original Eliza.~\cite{RadioLab}}
+
 % eliza interactions, show fragment
-\subsection{Web technologies}\label{sec:turtles}
+Each story module can include AIML fragments extending the
+conversational capabilities of Nell's agents.  Figure~\ref{fig:aiml}
+illustrates how this can be used to implement a traditional
+text adventure interaction for a particular story module.
+The child can write instructions such as, ``go north'', ``go
+east'', or ``open door'' for the agent.  These cause trigger events
+which the activity responds to to move the agent through the
+adventure---for example, by drawing the new location and causing
+the agent to describe it.
+
+% xx can the activity influence which commands are active?
+
+\subsection{TurtleScript}\label{sec:turtles}
+Nell's implementation leverages Web technologies, including HTML5,
+WebGL, and JavaScript, with extensive use of offline caching.
+Resources are named by URL, even when disconnected from the
+internet.  This allows third parties to manage their own namespaces
+when extending Nell, and simplifies updates to story modules and
+the Nell system itself.
+% if the kid's going to learn some software stack, this seem like a
+% fine one to learn
+
+As far as possible, Nell is implemented in JavaScript.  We have
+developed a block-oriented view of JavaScript, which we call
+TurtleScript~\cite{turtlescript}.  Combined with subsets of JavaScript
+for beginners, this allows us to use a single implementation language
+which scales from logo-like tasks all the way through the
+implementation of the entire Nell system itself.  TurtleScript's block
+view of JavaScript is isomorphic to the standard textual
+representation of code.  We have begun implementation of a full
+programming environment for TurtleScript which is naturally-suited to
+tablet machines.
+
+% XXX refer to TurtleScript excerpt in figure somewhere here.
+
+Our goal is to allow the child to eventually dig as far into Nell's
+implementation as their capabilities allow.  Although the operating system and
+the browser itself are inaccessible from the web platform, we can
+provide full access above that level.  The existence of meta-circular
+interpreters for JavaScript~\cite{narcissus} implies that we can
+expose the system all the way down to the programming language
+implementation.  Via TurtleScript, this code is all visible in a
+friendly block-based form.
+
 % Javascript/HTML5
@@
 \section{Related work}\label{sec:related}
+
 % Nell focused on Literacy

### 2012-03-09 paper.tex
Rough in related work and conclusions.

@@
 
-% Nell focused on Literacy
-% olpc context here?
-\cite{chang:tinkrbook}
-%context: english literacy, far from teachers & schools
-\subsection{The Sugar Learning Environment}
-% Sugar
-%% customization
-%% path through the educational material?
-\subsection{Narrative computing}
-% Reidl
-\subsection{Storytelling}
-% TinkRBooks
+The TinkRBook system~\cite{chang:tinkrbook} shares a similar focus on
+personalization and literacy.  Project Nell attempts to scale these
+ideas with a modular extensible narrative system and a broader range
+of pedagogic material.
+
+The Sugar Learning environment~\cite{XXX} is an earlier effort at
+child education.  We've learned directly from this effort.  Nell
+focuses on children even further from educational infrastructure.
+The narrative at Nell's code attempts to provide students additional
+pedagogical guidance, and the direct interaction model avoids keyboard
+problems which have plagued Sugar's initial deployments.  Nell also
+attempts to provide a greater sense of ownership through pervasive
+customization.
+
+Automatic narrative generation has seen attention in
+gaming, simulation, and training contexts~\cite{riedlXXX}.  Nell's
+use of these ideas in a pedagogical context is unique.
 
 \section{Conclusions}
+We have described the design of a novel narrative direct-interface
+system for teaching literacy, which is personalized for and grows with
+its child owner.  Personalized direct interface engages the child, and
+narrative guides them through Nell's pedagogic material.  The low-cost
+solar-powered hardware allows Nell to reach further into the
+least-developed areas of the world, to help those far from traditional
+educational infrastructure.
+
+We are in the process of implementing the Nell system, and expect to
+do initial deployments in Africa later this year.
 

### 2012-03-09 paper.tex
Tweak bibliography.

@@
 reasonable approximations do not have to be difficult.  We leverage
-the AIML language developed for AliceBot~\cite{xxx}, which has
+the AIML language developed for AliceBot~\cite{alicebotxxx}, which has
 won several Turing contests. \textbf{XXX more precise here XXX}
@@
 
-The Sugar Learning environment~\cite{XXX} is an earlier effort at
+The Sugar Learning environment~\cite{sugarXXX} is an earlier effort at
 child education.  We've learned directly from this effort.  Nell

### 2012-03-10 paper.tex
Citations for ELIZA.

@@
 \documentclass{acm_proc_article-sp}
+\usepackage{url}
 
@@
 Nell, directly.  Convincing discourse is a hard problem, but
-reasonable approximations do not have to be difficult.  We leverage
+reasonable approximations do not have to be difficult; even
+the rudimentary conversational abilities of ELIZA elicited hours of
+conversation~\cite{weizenbaum:power}.
+% her secretary is divulging her life story
+% asks him to leave the room so he could be alone with it
+% hours discussing her boyfriend.
+Nell's use as an interactive diary is enhanced by the child's
+collaboration in the fiction that Nell is intelligent---what
+Turkle~\cite{turkle:alone} calls the ``ELIZA effect.''
+
+\footnote{Wilcox identifies several shortcomings with the expressive
+  power of AIML, proposing ChatScript as a
+  replacement~\cite{wilcox:2010}.  At present AIML's community and
+  diversity of third-party implementations make it preferable.}
+
+  We leverage
 the AIML language developed for AliceBot~\cite{alicebotxxx}, which has
 won several Turing contests. \textbf{XXX more precise here XXX}
-\footnote{Anecdotal evidence shows that interactions with Eliza-style
-  responders can be rewarding: one woman spent hours each day
-  conversing with the original Eliza.~\cite{RadioLab}}
 
-% eliza interactions, show fragment
+% show fragment
 Each story module can include AIML fragments extending the
@@
 which the activity responds to to move the agent through the
-adventure---for example, by drawing the new location and causing
+adventure---for example, by drawing the new location on the page and causing
 the agent to describe it.

### 2012-03-10 paper.tex
Some AIML cites.

@@
 
+We leverage AIML~\cite{aiml:2005},%
 \footnote{Wilcox identifies several shortcomings with the expressive
@@
   diversity of third-party implementations make it preferable.}
-
-  We leverage
-the AIML language developed for AliceBot~\cite{alicebotxxx}, which has
-won several Turing contests. \textbf{XXX more precise here XXX}
+the markup language developed for AliceBot, three-time winner of the
+Loebner Prize, a restricted Turing Test competition for chat bots.
 

### 2012-03-10 paper.tex
AIML fragment.

@@
 \usepackage{url}
+%\usepackage{listings}
 
@@
 
-\subsection{Extensible dialog with AIML}
+\subsection{Extensible dialog}
 The child can interact with the Princess, and other agents within
@@
 
+\begin{figure}
+\begin{verbatim}
+<aiml version="1.0">
+<category>
+ <pattern>GO *</pattern>
+ <template><srai><star/></srai></template>
+</category>
+
+<category>
+ <pattern>N</pattern>
+ <template><srai>NORTH</srai></template>
+</category>
+
+<category>
+ <pattern>NORTH</pattern>
+ <template>
+  <javascript>FireEvent(Chat, "NORTH")</javascript>
+ </template>
+</category>
+</aiml>
+\end{verbatim}
+\caption{An AIML fragment which recognizes the commands ``N'',
+  ``North'', and ``Go North''.}\label{fig:aiml}
+\end{figure}
 % show fragment
@@
 text adventure interaction for a particular story module.
-The child can write instructions such as, ``go north'', ``go
-east'', or ``open door'' for the agent.  These cause trigger events
-which the activity responds to to move the agent through the
-adventure---for example, by drawing the new location on the page and causing
-the agent to describe it.
+The child can write instructions such as, ``go north,'' ``go
+east,'' or ``open door'' for the agent.  These trigger events
+handled by the story module to move the agent through the
+adventure---for example, by causing the activity to draw a
+new location, causing the agent to describe it, and shifting the
+AIML topic to enable vocabulary suitable for the new location.
 
 % xx can the activity influence which commands are active?
+% yes, by shifting the topic.  AIML should have a more flexible
+% topic system.
 

### 2012-03-10 paper.tex
First version of story model fragment.  Blows our page budget! (sigh)

@@
 \centering
-\epsfig{file=roger1.eps, height=1.5in} % width=3.2in
-\caption{Illustration of Nell's primary interface.}\label{fig:nell}
+\epsfig{file=xo3-3.eps, width=3.2in}
+\caption{Nell's solar-powered hardware platform.  The tablet design
+  allows for direct interaction.}\label{fig:xo3}
 \end{figure}
-
 % tell a nell story up front
@@
 authorship.
+
+\begin{figure}
+\centering
+\epsfig{file=roger1.eps, height=1.5in} % width=3.2in
+\caption{Illustration of Nell's primary interface.}\label{fig:nell}
+\end{figure}
 
@@
 
-\begin{figure}
-\centering
-\epsfig{file=xo3-3.eps, width=3.2in}
-\caption{Nell's solar-powered hardware platform.  The tablet design
-  allows for direct interaction.}\label{fig:xo3}
-\end{figure}
-
 The constructionist learning philosophy emphasizes creation and
@@
 \subsection{Rule-based story modules}
+\begin{figure*}
+\begin{verbatim}
+{ world: {
+    id: "/nell.laptop.org/chapters/alphabet-book" // canonical URL for this module
+    scene: "opening", // initial scene
+  }, actors: [
+    // these characters are common to different modules
+    { id: "user", type: "/nell.laptop.org/characters/user" },
+    { id: "princess", type: "/nell.laptop.org/characters/nell" }
+  ], scenes: [
+    { id: "opening", // http://nell.laptop.org/chapters/alphabet-book/scenes/opening
+      subject: "actors/princess", verb: "speak", object: "actors/user",
+      text: "Hello!  Would (o/2) like to hear a story?" },
+    { id: "story-1",
+      cond: rule`At(opening) && Event(Chat, Yes)`,
+      subject: "actors/princess", verb: "speak", object: "actors/user",
+      text: "Once upon a time..." },
+    ...
+    { id: "r-drawn",
+      cond: rule`Event(Activity, DrawR)`,
+      subject: "actors/princess", verb: "speak", object: "actors/user",
+      text: "Very good, (o/3)!",
+      action: js`startAnimation("r-runs")` },
+    ...
+  ]
+}
+\end{verbatim}
+\caption{Excerpt of JSON-formatted story module.  Some details (format
+  versioning, etc) have been elided).}\label{fig:rules}
+\end{figure*}
 Narrative is the core of the Nell interface.  It is important that
@@
 architected by Walter Bender, and through stimulating discussions with
-Chris Ball and Angela Chang.  Michael Stone also contributed to an
-early implementation.
+Chris Ball and Angela Chang.  Chris Ball and Michael Stone contributed
+to an implementation of the rule-based story model.
 

### 2012-03-10 paper.tex
Reformat and trim the story module example.

--- 2012-03-10 paper.tex
@@
 \subsection{Rule-based story modules}
+\begin{figure}\small
+\begin{verbatim}
+{ world: { // id gives canonical URL for this module
+    id: "/nell.laptop.org/chapters/alphabet-book"
+    scene: "opening", // initial scene
+  }, scenes: [
+    { id: "opening", subject: "actors/princess",
+      verb: "speak", object: "actors/user",
+      text: "Hello! Would (o/2) like to hear a story?"
+    },
+    { id: "story-1",
+      cond: rule`At(opening) && Event(Chat, Yes)`,
+      subject: "actors/princess",
+      verb: "speak", object: "actors/user",
+      text: "Once upon a time..." },
+    ...
+    { id: "r-drawn",
+      cond: rule`Event(Activity, DrawR)`,
+      subject: "actors/princess",
+      verb: "speak", object: "actors/user",
+      text: "Very good, (o/3)!",
+      action: js`startAnimation("r-runs")` },
+    ...
+  ], ... }
+\end{verbatim}
+\caption{Excerpt of story module.  JSON syntax has been extended with
+quasiquote~\cite{quasiquote}.}\label{fig:rules}
+\end{figure}
 Narrative is the core of the Nell interface.  It is important that
@@
 
-\begin{figure}
+\begin{figure}\small
 \begin{verbatim}
@@
  <template>
-  <javascript>FireEvent(Chat, "NORTH")</javascript>
+  <javascript>FireEvent(Chat, "North")</javascript>
  </template>

### 2012-03-10 paper.tex
Gratuitously inherit from a prototypical drawing activity.

@@
 { world: { // id gives canonical URL for this module
-    id: "/nell.laptop.org/chapters/alphabet-book"
+    id: "/nell.laptop.org/chapters/alphabet-book",
     scene: "opening", // initial scene
+    inherit_from: "/nell.laptop.org/core/drawing_activity",
   }, scenes: [

### 2012-03-10 paper.tex
Story module citations.

@@
 
-XXX stories consist of elements.  all elements contain preconditions,
-and actions.  cite PLANNER.
+Stories consist of a sequence of \textit{scenes}, in addition to
+characters, locations, and activity information.  Scenes are ``goals''
+in the STRIPS formalism~\cite{strips}, and consist of
+action descriptions, preconditions, and effects.
+In principle, all preconditions for all scenes of all story modules
+are evaluated continuously to determine the actions of the narrative
+agents.  This is made computationally efficient using the Rete
+algorithm.  Dialog is expressed using Curveship-style
+templates~\cite{montfort:curveship}.
 
-In principle, all rules for all elements of all stories are evaluated
-continuously to determine the actions of the narrative agents.  This
-is made computationally efficient using the Rete
-algorithm~\cite{rete}.  The Rete algorithm requires that our
-conditions are in Normal Form\footnote{BE PRECISE HERE}, but actions
-can be arbitrarily complex.
+%  The Rete algorithm requires that our conditions are in Normal
+%  Form\footnote{BE PRECISE HERE}, but actions can be arbitrarily complex.
 
@@
 
-Story modules can be inherited and extended, as figure~\ref{fig:rules2}
-shows.  This makes it easy to take an existing story and modify it to
+Story modules can be inherited and extended, as shown in the figure.
+This makes it easy to take an existing story and modify it to
 better suit particular interests or a particular type of learner, or

### 2012-03-11 paper.tex
Note feedback from mstone around the relevant parts of the paper.

@@
 
+% MSTONE FEEDBACK: lead off w/ responding to learning styles, not
+% changing colors.  (Also incorporate learning style stuff into
+% section 1.)  Move color changing into Related Work, even?
+
 Children often decorate their belongings, and the
@@
 \subsection{Growable}
+
+% MSTONE FEEDBACK: 'accumulating' in 3rd sentence starts conflating
+% serial 'tech tree' of story material w/ 'decentralized authoring'
+% mechanism of expanding topics covered by the system.
+% Move 'tech tree' aspect to personalization: ``ie, not only does
+% the system personalize based on learning style, but also based
+% on achievement''.  last paragraph can segue into 'growable'
+% which is mostly about 3rd party expansion.
+
 The Nell system aims to have a ``low floor and no ceiling.''  The low
@@
 
+% MSTONE FEEDBACK: ``the innovations vis-a-vis Sugar are: a) authoring
+% and publishing is a first-class concept, instead of just
+% "synchronous collaboration" and b) there's a pedagogically motivated
+% feedback loop; it's not just an open-loop "constructionist" system''
+
 Automatic narrative generation has seen attention in
@@
 Chris Ball and Angela Chang.  Chris Ball and Michael Stone contributed
-to an implementation of the rule-based story model.
+to an implementation of the rule-based story model, and Michael Stone
+provided valuable feedback on this paper.
 

### 2012-03-11 paper.tex
Describe rule group mechanism.

@@
       verb: "speak", object: "actors/user",
-      text: "Hello! Would (o/2) like to hear a story?"
+      text: "Hello! Would (o/2) like to hear a story?",
+      group: "/nell.laptop.org/core/story-intro-group",
     },
@@
 are evaluated continuously to determine the actions of the narrative
-agents.  This is made computationally efficient using the Rete
+agents.  This production rules system
+is made computationally efficient using the Rete
 algorithm.  Dialog is expressed using Curveship-style
@@
 just to add variety.  This can lead to conflicts: multiple versions of
-a story, or multiple variant elements within a story, may have their
+a story, or variant scenes within a story, may have their
 preconditions satisfied simultaneously.
 
-We resolve conflicts...
+Conflicts are resolved by consulting a \textit{group} associated with
+each scene.  Each group has a default priority and names a scene to be
+invoked when a conflict exists; the actions associated with that scene
+will temporarily alter the priorities of the conflicting scenes to
+resolve the conflict.  For example, the \textit{story-intro-group}
+cited in the example defaults to a lower priority, such that
+continuing a story in progress is preferred to beginning a new scene.
+The resolution scene for \textit{story-intro-group} may be part of its
+own story: for example, it may bring us to the controls of a magic airship
+which allows us to pilot between locations corresponding to the
+various story continuations.  Selecting a destination location bumps
+the priority of the given scene allowing us to transition to it
+without further conflict.
+
+Other rule groups may invoke random selection, select between stories
+based on child preferences or curriculum, or (during development) flag
+an error to highlight an unexpected ambiguity.
 
@@
 \caption{An AIML fragment which recognizes the commands ``N'',
-  ``North'', and ``Go North''.}\label{fig:aiml}
+  ``North'', and ``Go North''. The \verb|<srai>| tag invokes the
+  recognizer recursively.}\label{fig:aiml}
 \end{figure}

### 2012-03-11 paper.tex
Fix broken caption.

@@
 \end{verbatim}
-\caption{An AIML fragment which recognizes the commands ``N'',
-  ``North'', and ``Go North''. The \verb|<srai>| tag invokes the
+\caption{An AIML fragment which recognizes the commands \texttt{N},
+  \texttt{North}, and \texttt{Go North}. The \texttt{<srai>} tag invokes the
   recognizer recursively.}\label{fig:aiml}

### 2012-03-11 paper.tex
Add lots of co-authors!

@@
 
-\numberofauthors{1}
+\numberofauthors{3}
 \author{
@@
 \email{cscott@laptop.org}
+% 2nd. author
+\alignauthor
+Chris Ball\\
+\affaddr{One Laptop Per Child Foundation}\\
+\affaddr{222 Third Street}\\
+\affaddr{Cambridge, MA 02142}\\
+\email{cjb@laptop.org}
+% 3rd. author
+\alignauthor
+Michael Stone\\
+\affaddr{x}\\
+% michael, fill in whatever affiliation you like here.
+\email{mstone@laptop.org} % or other email address
 %\and  % use '\and' if you need 'another row' of author names
@@
 architected by Walter Bender, and through stimulating discussions with
-Chris Ball and Angela Chang.  Chris Ball and Michael Stone contributed
-to an implementation of the rule-based story model, and Michael Stone
-provided valuable feedback on this paper.
-
+the Sugar community and Angela Chang.
 %

### 2012-03-11 paper.tex
Rewrite abstract.

@@
 % ONE 'self-contained' source file.
-\documentclass{acm_proc_article-sp}
+%\documentclass{acm_proc_article-sp}
+\documentclass[preprint]{sig-alternate}
 \usepackage{url}
-%\usepackage{listings}
+\usepackage{ifthen}\newboolean{noauthors}\setboolean{noauthors}{false}
 
-\begin{document}
+\begin{document}\toappear{}
 
@@
 
-\numberofauthors{3}
+\ifthenelse{\boolean{noauthors}}{}{
+\numberofauthors{1}
 \author{
@@
 \alignauthor
-C. Scott Ananian\\
+C. Scott Ananian \qquad Chris Ball \qquad Michael Stone\titlenote{%
+Now at Akamai Technologies.}\\
 \affaddr{One Laptop Per Child Foundation}\\
@@
 \affaddr{Cambridge, MA 02142}\\
-\email{cscott@laptop.org}
-% 2nd. author
-\alignauthor
-Chris Ball\\
-\affaddr{One Laptop Per Child Foundation}\\
-\affaddr{222 Third Street}\\
-\affaddr{Cambridge, MA 02142}\\
-\email{cjb@laptop.org}
-% 3rd. author
-\alignauthor
-Michael Stone\\
-\affaddr{x}\\
-% michael, fill in whatever affiliation you like here.
-\email{mstone@laptop.org} % or other email address
+\email{\{cscott,cjb,mstone\}@laptop.org}
 %\and  % use '\and' if you need 'another row' of author names
+}
 }
@@
 \begin{abstract}
-Project Nell is tablet software to teach reading and writing to
-children far from educational infrastructure.  Nell is a modular
-narrative system with ``a low floor and no ceilings'': the design
-scales from teaching letter shapes to programming the system itself.
+% Draft 1
+% Project Nell is tablet software to teach reading and writing to
+% children far from educational infrastructure.  Nell is a modular
+% narrative system with ``a low floor and no ceilings'': the design
+% scales from teaching letter shapes to programming the system itself.
+% Nell is the successor to the Sugar educational software developed by
+% One Laptop per Child for their XO-1 laptop.  Nell targets the XO-3
+% tablet, a sub-\$100 solar-powered tablet for the developing world.
+% %Experience with Sugar drove design choices for Nell.
 
-Nell is the successor to the Sugar educational software developed by
-One Laptop per Child for their XO-1 laptop.  Nell targets the XO-3
-tablet, a sub-\$100 solar-powered tablet for the developing world.
-%Experience with Sugar drove design choices for Nell.
+% Draft 2
+% Motivated by ``lessons learned'' from five years of experience with our
+% \textit{Sugar} education platform, we introduce \textit{Nell}: a Web- and
+% tablet\footnote{Nell targets the XO-3 tablet, a sub-\$100 solar-powered tablet
+% for the developing world.}-friendly education platform built around a novel
+% modular narrative system designed to engage children far from educational
+% infrastructure in their own personal pursuit of education and literacy.
 
-\textbf{XXX REWRITE THE ABSTRACT XXX}
+% Draft 3
+% We introduce \textit{Nell}: a Web- and tablet\footnote{Nell targets the XO-3
+% tablet, a sub-\$100 solar-powered tablet for the developing world.}-friendly
+% education platform built around a novel modular narrative system designed to
+% engage children far from educational infrastructure in their own personal
+% pursuit of education and literacy. Our design is informed and motivated by
+% ``lessons learned'' from five years and $\sym2,000,000$ users' experience with
+% our \textit{Sugar} education platform.
+
+% Draft 4
+% \textit{Nell} is a tablet-friendly\footnote{Nell's primary platform is the
+% XO-3, an affordable solar-powered tablet for kids in the developing world.}
+% education platform with a novel modular narrative system designed to engage
+% children far from educational infrastructure in their own personal pursuit of
+% literacy. Nell's design is informed by ``lessons learned'' from five years'
+% experience with the first-generation \textit{Sugar} education platform,
+% deployed to two million users.
+
+% Draft 5
+% \textit{Nell} is a tablet-friendly\footnote{Nell's primary platform is the
+% XO-3, an affordable solar-powered tablet for kids in the developing world.}
+% education platform with a novel modular narrative system designed to engage
+% children in their own personal exploration of the world around them. Nell
+% extends the design of the \textit{Sugar Learning
+% Platform}\footnote{\url{http://sugarlabs.org}} to support new features
+% requested by the field: \textit{user-editable scaffolding},
+% \textit{personalized learning materials}, and \textit{improved authoring
+% workflows}.
+
+% Draft 6
+\textit{Nell} is a tablet-friendly\footnote{Nell's primary platform is the
+XO-3, an affordable solar-powered tablet for kids in the developing world.}
+education platform with a novel modular narrative system designed to engage
+children in their own personal pursuit of literacy. Nell extends the design of
+the \textit{Sugar Learning Platform}\footnote{\url{http://sugarlabs.org}} to
+better support \textit{narrative} and \textit{personalization} and to better
+accommodate \textit{growth}, \textit{authoring}, and \textit{direct
+interaction}.
 \end{abstract}
@@
 
-We have not yet implemented an integrated story editor within Nell,
-but we expect it to be roughly similar to the one implemented for
-Wide Ruled, which showed good learnability by inexperienced
-users~\cite{wideruled}.
+Nell's integrated story editor has not yet been implemented.  Its
+design is patterned after the Wide Ruled authoring tool, which was
+successfully used by non-technical story creators~\cite{skorupski:2009}.
 
@@
 \subsection{Extensible dialog}
-The child can interact with the Princess, and other agents within
-Nell, directly.  Convincing discourse is a hard problem, but
+The child can directly interact with the Princess and other agents within
+Nell.  Convincing discourse is a hard problem, but
 reasonable approximations do not have to be difficult; even
@@
 
-The Sugar Learning environment~\cite{sugarXXX} is an earlier effort at
+The Sugar Learning Platform~\cite{sugar} is an earlier effort at
 child education.  We've learned directly from this effort.  Nell

### 2012-03-11 paper.tex
Switch to unified affiliation.
Add Wide Ruled citation, clean up reference to it.

--- 2012-03-11 paper.tex
@@
 
-\numberofauthors{3}
+\numberofauthors{1}
 \author{
@@
 \affaddr{Cambridge, MA 02142}\\
-\email{cscott@laptop.org}
-% 2nd. author
-\alignauthor
-Chris Ball\\
-\affaddr{One Laptop Per Child Foundation}\\
-\affaddr{222 Third Street}\\
-\affaddr{Cambridge, MA 02142}\\
-\email{cjb@laptop.org}
-% 3rd. author
-\alignauthor
-Michael Stone\\
-\affaddr{x}\\
-% michael, fill in whatever affiliation you like here.
-\email{mstone@laptop.org} % or other email address
+\email{\{cscott,cjb,mstone\}@laptop.org}
 %\and  % use '\and' if you need 'another row' of author names

### 2012-03-11 paper.tex
Cite Sugar properly (?).
Switch to tighter "alternate" layout.  Much closer to our page budget now!
Add mechanism to suppress authors for review copy.
Properly cite Mark Riedl paper.
Tweak section; remove the Raven and use a flag rather than boots.

--- 2012-03-11 paper.tex
@@
 % tell a nell story up front
-Miles from the nearest school, a young Ethiopian girl turns on a
-tablet computer for the first time.  The solar-powered machine begins
-talking to her immediately: \textit{``Hello!  Would you like to hear a story?''}
+Miles from the nearest school, a young Ethiopian girl named Rahel turns on her
+new tablet computer. The solar-powered machine greets her:
+\textit{``Hello!  Would you like to hear a story?''}
 
 She nods and listens to a story about a princess.  Later, when
-the girl has learned a little more, she will tell the machine that the
-princess is named ``Rahel,'' like she is, and change all its
-colors---but for now the green book draws pictures of the unnamed Princess
-for her and asks her to trace shapes on the screen.
+the Rahel has learned a little more, she will tell the machine that the
+princess is named ``Rahel'' like she is and she will change the princess'
+colors---but for now the green book draws pictures of the unnamed Princess for
+her and asks her to trace shapes on the screen.
 
-\textit{``R is for Raven.  Can you trace the R?''}
+\textit{``R is for Run.  Can you trace the R?''}
 
 As she traces the R, it comes to life and gallops across the screen.
-\textit{``Run starts with R.  Roger the R runs across the Red Rug.  Roger has a
-dog named Rover.''}  Rover barks: \textit{``Ruff! Ruff!''}  The Princess asks,
-\textit{``Can you find something Red?''} and Rahel uses the camera to
-photograph her new boots.
+\textit{``Run starts with R.  Roger the R runs across the Red Rug.
+Roger has a dog named Rover.''}  Rover barks: \textit{``Ruff! Ruff!''}  The
+Princess asks, \textit{``Can you find something Red?''} and Rahel uses the
+camera to photograph the red stripe on a nearby Ethiopian flag.
 
@@
 Automatic narrative generation has seen attention in
-gaming, simulation, and training contexts~\cite{riedlXXX}.  Nell's
+gaming, simulation, and training contexts~\cite{riedl:2008}.  Nell's
 use of these ideas in a pedagogical context is unique.

### 2012-03-11 paper.tex
Polish Michael's polishing of section 1; add some Multiple Intelligences stuff.

@@
 Miles from the nearest school, a young Ethiopian girl named Rahel turns on her
-new tablet computer. The solar-powered machine greets her:
+new tablet computer. The solar-powered machine speaks to her:
 \textit{``Hello!  Would you like to hear a story?''}
@@
 the Rahel has learned a little more, she will tell the machine that the
-princess is named ``Rahel'' like she is and she will change the princess'
-colors---but for now the green book draws pictures of the unnamed Princess for
+princess is named ``Rahel'' like she is, and that she likes to wear blue---but
+for now the green book draws pictures of the unnamed Princess for
 her and asks her to trace shapes on the screen.
@@
 Princess asks, \textit{``Can you find something Red?''} and Rahel uses the
-camera to photograph the red stripe on a nearby Ethiopian flag.
+camera to photograph berries on a nearby bush.
 
@@
 to draw different shapes).  Rahel writes her own stories about the Princess,
-which she shares with her friends.
+which she shares with her friends.  The book tells her that she is
+very good at music, and her lessons begin to encourage her to invent
+silly songs about what she's learning.
 
@@
 the turtle is also used to write all the software running inside
-her special book.  The stories Rahel writes can now include new
-functionality that make her book even more exciting for the kids
-just now receiving their own green meta books.
+her special book.  Rahel begins to write stories with different
+kinds of new music functionality, which make the green book even more
+exciting for her younger brother and the other younger kids just now
+receiving them.
 

### 2012-03-11 paper.tex
Abstracts are hard.

@@
 
-% Draft 6
-\textit{Nell} is a tablet-friendly\footnote{Nell's primary platform is the
-XO-3, an affordable solar-powered tablet for kids in the developing world.}
-education platform with a novel modular narrative system designed to engage
-children in their own personal pursuit of literacy. Nell extends the design of
-the \textit{Sugar Learning Platform}\footnote{\url{http://sugarlabs.org}} to
-better support \textit{narrative} and \textit{personalization} and to better
-accommodate \textit{growth}, \textit{authoring}, and \textit{direct
-interaction}.
+%% % Draft 6
+%% \textit{Nell} is a tablet-friendly\footnote{Nell's primary platform is the
+%% XO-3, an affordable solar-powered tablet for kids in the developing world.}
+%% education platform with a novel modular narrative system designed to engage
+%% children in their own personal pursuit of literacy. Nell extends the design of
+%% the \textit{Sugar Learning Platform}\footnote{\url{http://sugarlabs.org}} to
+%% better support \textit{narrative} and \textit{personalization} and to better
+%% accommodate \textit{growth}, \textit{authoring}, and \textit{direct
+%% interaction}.
+
+% Draft 7
+% "education platform" -> "literacy platform"?
+\textsc{Nell} is a tablet-oriented education platform for children in
+the developing world.  A novel modular narrative system guides
+learning, even for children far from educational infrastructure, and
+provides personalized instruction which grows with the child.
+Nell's design builds on experience with the Sugar Learning
+Platform~\cite{sugar}, used by over a million children around the
+world.
+
 \end{abstract}

### 2012-03-11 paper.tex
Misc. tweaks for space.

@@
   Uses in Education}[computer-managed instruction]
-\category{I.2.7}{Artificial Intelligence}{Natural Language
-  Processing}[narrative interfaces]
+%\category{I.2.7}{Artificial Intelligence}{Natural Language
+%  Processing}[narrative interfaces]
 \category{H.5.2}{Information Interfaces and Presentation}{User
@@
 stories~\cite{boyd:stories}.  Nell uses a
-\textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative,laurel:computers} 
+\textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}
 to guide the child user through the pedagogical capabilities of the
@@
 show how a planner can be used to recombine and adapt story fragments.
-In this we expect to be less ambitious than the cited work: instead of
+We have less ambition than the cited work: instead of
 attempting to generate thousands of stories from tens of templates, we
@@
       text: "Very good, (o/3)!",
-      action: js`startAnimation("r-runs")` },
-    ...
+      action: js`startAnimation("r-runs")` }, ...
   ], ... }
@@
 
-We leverage AIML~\cite{aiml:2005},%
-\footnote{Wilcox identifies several shortcomings with the expressive
-  power of AIML, proposing ChatScript as a
-  replacement~\cite{wilcox:2010}.  At present AIML's community and
-  diversity of third-party implementations make it preferable.}
-the markup language developed for AliceBot, three-time winner of the
-Loebner Prize, a restricted Turing Test competition for chat bots.
-
 \begin{figure}\small
@@
 \end{figure}
-% show fragment
+
+We leverage AIML~\cite{aiml:2005},
+the markup language developed for the three-time Loebner Prize-winning
+AliceBot.%
+\footnote{Wilcox identifies several shortcomings with the expressive
+  power of AIML, proposing ChatScript as a
+  replacement~\cite{wilcox:2010}.  At present AIML's community and
+  diversity of third-party implementations make it preferable.}
 Each story module can include AIML fragments extending the
-conversational capabilities of Nell's agents.  Figure~\ref{fig:aiml}
-illustrates how this can be used to implement a traditional
+conversational capabilities of Nell's agents.
+
+Figure~\ref{fig:aiml}
+shows how this can be used to implement a traditional
 text adventure interaction for a particular story module.
-The child can write instructions such as, ``go north,'' ``go
+In this example the child can write instructions such as, ``go north,'' ``go
 east,'' or ``open door'' for the agent.  These trigger events
@@
 
-\section{Acknowledgments}
-Nell's design evolved from experience with the Sugar system,
-architected by Walter Bender, and through stimulating discussions with
-the Sugar community and Angela Chang.
+% CSA: Acknowledge the below in the related work, rather than spending
+% space on a heading and a new section.
+
+%% \section{Acknowledgments}
+%% Nell's design evolved from experience with the Sugar system,
+%% architected by Walter Bender, and through stimulating discussions with
+%% the Sugar community and Angela Chang.
+
 %

### 2012-03-11 paper.tex
More tweaks for space.  References now take up only 1 col (12.5% total space).

@@
 \subsection{Narrative}
+%used to \cite{flores:uruguay,idb:peru} below; removed for space
 
 Experience with our previous educational software
-system~\cite{flores:uruguay,idb:peru} has shown gratifying uptake and
+system has shown gratifying uptake and
 use by children.  But we are often disappointed that some of our
@@
 cues from \textit{The Diamond Age} and
-from the UNIVERSE narrative-generation system \cite{lebowitz:universe84}.
+from the UNIVERSE narrative-generation system \cite{lebowitz:universe85}.
 Each of the several characters represents a specific skill or subject
@@
 \epsfig{file=roger1.eps, height=1.5in} % width=3.2in
-\caption{Illustration of Nell's primary interface.}\label{fig:nell}
+\caption{Diagram of Nell's primary interface.}\label{fig:nell}
 \end{figure}
@@
 can engage the child's interests and cater to different learning
-styles~\cite{gardner:mi}.
+styles.%~\cite{gardner:mi}.%%%%%%% <- lead with this.
 
@@
   power of AIML, proposing ChatScript as a
-  replacement~\cite{wilcox:2010}.  At present AIML's community and
-  diversity of third-party implementations make it preferable.}
+  replacement~\cite{wilcox:2010}.  AIML's community and
+  diversity of third-party implementations make it preferable at present.}
 Each story module can include AIML fragments extending the
@@
 The TinkRBook system~\cite{chang:tinkrbook} shares a similar focus on
-personalization and literacy.  Project Nell attempts to scale these
+personalization and literacy.  Nell attempts to scale these
 ideas with a modular extensible narrative system and a broader range

### 2012-03-11 paper.tex
Add a citation for HMM handwriting recognition

@@
 implementation of a separated-character handwriting recognition engine
-using Hidden Markov Models is well-understood.
+using Hidden Markov Models is well-understood~\cite{hanshu:handwriting}.
 \textbf{SMOOTH THIS LANGUAGE}

### 2012-03-11 paper.tex
Add a missing ~ before \cite{}

@@
 cues from \textit{The Diamond Age} and
-from the UNIVERSE narrative-generation system \cite{lebowitz:universe85}.
+from the UNIVERSE narrative-generation system~\cite{lebowitz:universe85}.
 Each of the several characters represents a specific skill or subject

### 2012-03-11 paper.tex
over one million children -> over two million
We can use http://wiki.laptop.org/go/Deployments as a citation -- it
says "There are over 2.5 million XOs in the field as of November, 2011."
Some of those are in warehouses, but not .5 million of them.
Typo fix: "In a disconnected deployments"

@@
 Nell's design builds on experience with the Sugar Learning
-Platform~\cite{sugar}, used by over a million children around the
+Platform~\cite{sugar}, used by over two million children around the
 world.
@@
 New story modules can be incorporated by visiting an internet page in
-connected deployments.  In a disconnected deployments, new stories
+connected deployments.  In disconnected deployments, new stories
 might be distributed once a year on USB sticks, or shared from your

### 2012-03-11 paper.tex
Add a conflict resolution example of verbal vs. visual learning style

@@
 Other rule groups may invoke random selection, select between stories
-based on child preferences or curriculum, or (during development) flag
-an error to highlight an unexpected ambiguity.
+based on child preferences or curriculum (for example, the system
+could attempt to learn a child's preference for verbal vs. visual
+learning), or (during development) flag an error to highlight an
+unexpected ambiguity.
 

### 2012-03-11 paper.tex
Tweak punctuation in S.1.

@@
 the Rahel has learned a little more, she will tell the machine that the
-princess is named ``Rahel'' like she is, and that she likes to wear blue---but
+princess is named ``Rahel'' like she is and that she likes to wear blue---but
 for now the green book draws pictures of the unnamed Princess for

### 2012-03-11 paper.tex
Simplify language in the "older-Rahel" story.

@@
 An older Rahel learns that the block language she used to talk with
-the turtle is also used to write all the software running inside
-her special book.  Rahel begins to write stories with different
-kinds of new music functionality, which make the green book even more
+the turtle is also used to write all the software and stories running inside
+her special book.  Rahel begins to write new stories with her own
+music and dialog, which make the green book even more
 exciting for her younger brother and the other younger kids just now

### 2012-03-11 paper.tex
Tighten a few Technology sentences.

@@
 
-\section{A Lesson from Nell}
+\section{A Lesson from Nell} \label{sec:lesson}
 \begin{figure}
@@
 \section{Technology}
-We have begun implementation of the Nell system.  In this section we
-will describe the implementation of three of the most novel pieces
-from the software engineering side.  These are not necessarily the
-most novel-appearing from the user's perspective: for example, the
-handwriting UI is still fairly unusual in consumer devices, but the
-implementation of a separated-character handwriting recognition engine
-using Hidden Markov Models is well-understood~\cite{hanshu:handwriting}.
-\textbf{SMOOTH THIS LANGUAGE}
+We describe the novel synthesis of technologies that makes up our current Nell
+prototype. Several of these technologies are well-established; a few (like
+``TurtleScript'') are novel.
+
+\subsection{Handwriting recognition}
+To support \textit{direct interaction}, we use a separated-character
+handwriting recognition engine based on Hidden Markov Models~\cite{hanshu:handwriting}.
 
@@
 \end{figure}
-Narrative is the core of the Nell interface.  It is important that
-Nell's story be modular and extensible.  We use a rule-based story
-model with programmable conflict resolution to provide this ability.
-Figure~\ref{fig:rules} shows an excerpt of the story module driving
-the interaction described in the introduction to this paper.
+To support Nell's \textit{modular narrative interface}, we use a rule-based
+story model with programmable conflict resolution. (Figure~\ref{fig:rules} shows
+an excerpt of the story module underlying the interaction described in
+Section~\ref{sec:lesson}.)
 
-Stories consist of a sequence of \textit{scenes}, in addition to
-characters, locations, and activity information.  Scenes are ``goals''
-in the STRIPS formalism~\cite{strips}, and consist of
+Stories consist of a sequence of \textit{scenes} plus information about
+characters, locations, and activities.  Scenes are ``goals''
+in the STRIPS formalism~\cite{strips} and consist of
 action descriptions, preconditions, and effects.
@@
 connected deployments.  In disconnected deployments, new stories
-might be distributed once a year on USB sticks, or shared from your
+might be distributed once a year on USB sticks, or shared among
 friends.  In a classroom environment, a teacher can share stories for

### 2012-03-11 paper.tex
Abbreviate figure references.

@@
 computer to maintain the child's connection to their work
-(Figure~\ref{fig:xo3}).
+(Fig.~\ref{fig:xo3}).
 
@@
 To support Nell's \textit{modular narrative interface}, we use a rule-based
-story model with programmable conflict resolution. (Figure~\ref{fig:rules} shows
+story model with programmable conflict resolution. (Fig.~\ref{fig:rules} shows
 an excerpt of the story module underlying the interaction described in
@@
 resolve the conflict.  For example, the \textit{story-intro-group}
-cited in the example defaults to a lower priority, such that
+used in Fig.~\ref{fig:rules} defaults to a lower priority, such that
 continuing a story in progress is preferred to beginning a new scene.
@@
 
-Figure~\ref{fig:aiml}
+Fig.~\ref{fig:aiml}
 shows how this can be used to implement a traditional

### 2012-03-11 paper.tex
Tweak technology language; prefer "Web" to "internet".

@@
 
-New story modules can be incorporated by visiting an internet page in
+New story modules can be downloaded from the Web in
 connected deployments.  In disconnected deployments, new stories
-might be distributed once a year on USB sticks, or shared among
+might be distributed once a year on USB sticks or shared among
 friends.  In a classroom environment, a teacher can share stories for

### 2012-03-11 paper.tex
Streamline cjb's learning style example.

@@
 Other rule groups may invoke random selection, select between stories
-based on child preferences or curriculum (for example, the system
-could attempt to learn a child's preference for verbal vs. visual
-learning), or (during development) flag an error to highlight an
+based on child preferences, curriculum, or inferred learning style,
+or (during development) flag an error to highlight an
 unexpected ambiguity.

### 2012-03-11 paper.tex
Revert "Simplify language in the "older-Rahel" story."
This reverts commit b63497cd0917142a8afc8cf0bdbc20cf7eee4460.

I was trying to distinguish between writing stories (which even younger
Rahel was doing) and writing code (which is what older Rahel is learning
to do).  This paragraph could be clarified, but not in the direction of
the reverted commit.

@@
 An older Rahel learns that the block language she used to talk with
-the turtle is also used to write all the software and stories running inside
-her special book.  Rahel begins to write new stories with her own
-music and dialog, which make the green book even more
+the turtle is also used to write all the software running inside
+her special book.  Rahel begins to write stories with different
+kinds of new music functionality, which make the green book even more
 exciting for her younger brother and the other younger kids just now

### 2012-03-11 paper.tex
Remove 'during development' part of rule groups text.

@@
 
-Other rule groups may invoke random selection, select between stories
-based on child preferences, curriculum, or inferred learning style,
-or (during development) flag an error to highlight an
-unexpected ambiguity.
+Other rule groups may invoke random selection or select between stories
+based on child preferences, curriculum, or inferred learning style.
 

### 2012-03-11 paper.tex
Simplify key ideas intro.

@@
 and is \textbf{Personalized} for, the child.
-
-We believe the combination of these four key concepts provides a novel
-experience which addresses some of the shortcomings of earlier
-learning environments for children.  In this section we will discuss
-these four building blocks, and in the next section we will describe their
-implementation in Nell.
+Together, these four key concepts provide a novel
+experience which we use to address several challenges faced by earlier
+learning environments for children.
 

### 2012-03-11 paper.tex
Rewrite the intros in section 2.

@@
 
-Experience with our previous educational software
-system has shown gratifying uptake and
-use by children.  But we are often disappointed that some of our
-favorite pedagogical tools included in the systems were not being
-widely used.  Enthusiastic teachers would spark intensive use in
-certain classrooms, but children had difficulty finding their way into
-the material without strong guidance.
-
 Children (like all humans) are hard-wired for
-stories~\cite{boyd:stories}.  Nell uses a
-\textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}
-to guide the child user through the pedagogical capabilities of the
-system.  All system actions are shaped by the storybook metaphor, and
-interactions with the system are conceptualized as interactions with
-one of the system protagonists.
+stories~\cite{boyd:stories}, yet our prior Sugar Learning Platform
+made no use of this fact. By contrast, Nell augments Sugar's
+affordances for \textit{action}, \textit{reflection}, and
+\textit{communication} with new affordances for \textit{engagement} organized
+as a modular \textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}.
+In this interface, all system actions are shaped by the storybook metaphor, and
+interactions with the system are conceptualized as interactions with one of the
+system protagonists.
 
@@
 
-As illustrated in figure~\ref{fig:nell}, Nell's characters are
+As depicted in Figure~\ref{fig:nell}, Nell's characters are
 always-available agents layered above a particular system \textit{activity}
@@
 
-The Nell system aims to have a ``low floor and no ceiling.''  The low
+Children grow. To accommodate them, Nell provides a ``low floor and
+no ceiling.''  The low
 floor is a very simple base system, usable with no literacy skills or
 external guidance.  The system then grows with the child's expanding
-capabilities, serially accumulating story modules to tackle topics of
-increasing difficulty.  As mentioned before, the ability for third
-parties to independently author story modules is necessary to ensure
-Nell can continue to grow.  There should be no external dependencies
+capabilities, serially exposing story modules to tackle topics of
+increasing difficulty.  Later, the ability for third
+parties to independently author story modules ensures that
+Nell can continue to grow.  Finally, there are no external dependencies
 limiting a child's ability to participate in this process: a Nell
@@
 
-The constructionist learning philosophy emphasizes creation and
-tangible interaction.  Nell uses direct interaction on a tablet
-computer to maintain the child's connection to their work
+Children construct knowledge from direct interactive
+experience~\cite{Papert91}. As a result, (and like its predecessor Sugar), Nell
+emphasizes creation and tangible interaction by using direct interaction; this
+time, using a tablet computer to maintain the child's connection to their work
 (Fig.~\ref{fig:xo3}).

### 2012-03-11 paper.tex
Tone down sugar references in section 2.1.

@@
 Children (like all humans) are hard-wired for
-stories~\cite{boyd:stories}, yet our prior Sugar Learning Platform
-made no use of this fact. By contrast, Nell augments Sugar's
-affordances for \textit{action}, \textit{reflection}, and
-\textit{communication} with new affordances for \textit{engagement} organized
-as a modular \textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}.
+stories~\cite{boyd:stories}. Nell takes advantage of this fact 
+to augment prior affordances for \textit{action}, \textit{reflection}, and
+\textit{communication} with new affordances for \textit{engagement} in the form of
+a modular \textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}.
 In this interface, all system actions are shaped by the storybook metaphor, and

### 2012-03-11 paper.tex
Try to clarify the relationship between direct interaction, Nell, and constructionism.

@@
 
-Children construct knowledge from direct interactive
-experience~\cite{Papert91}. As a result, (and like its predecessor Sugar), Nell
-emphasizes creation and tangible interaction by using direct interaction; this
-time, using a tablet computer to maintain the child's connection to their work
-(Fig.~\ref{fig:xo3}).
+Children often learn from direct interactive experience. Consequently, Nell
+intensifies our commitment to constructionism~\cite{Papert91} by further
+emphasizing creation and tangible interaction using a tablet computer as a way
+to maintain the child's connection to their work (Fig.~\ref{fig:xo3}).
 

### 2012-03-11 paper.tex
Undo some of mstone's parallel structure, sorry.

@@
 
+Experience with our previous educational software
+system has shown gratifying uptake and
+use by children.  But we are often disappointed that some of our
+favorite pedagogical tools included in the systems were not being
+widely used.  Enthusiastic teachers would spark intensive use in
+certain classrooms, but children had difficulty finding their way into
+the material without strong guidance.
+
 Children (like all humans) are hard-wired for
-stories~\cite{boyd:stories}. Nell takes advantage of this fact 
-to augment prior affordances for \textit{action}, \textit{reflection}, and
-\textit{communication} with new affordances for \textit{engagement} in the form of
-a modular \textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}.
-In this interface, all system actions are shaped by the storybook metaphor, and
+stories~\cite{boyd:stories}.  Nell uses a
+\textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}
+to guide the child user through the pedagogical capabilities of the
+system. 
+In this interface, all system actions are shaped by a storybook metaphor, and
 interactions with the system are conceptualized as interactions with one of the
@@
 floor is a very simple base system, usable with no literacy skills or
-external guidance.  The system then grows with the child's expanding
-capabilities, serially exposing story modules to tackle topics of
-increasing difficulty.  Later, the ability for third
+external guidance.  The serial progression of story modules grows with
+the child to tackle topics of increasing difficulty.
+Later, the ability for third
 parties to independently author story modules ensures that
@@
 
-Children often learn from direct interactive experience. Consequently, Nell
-intensifies our commitment to constructionism~\cite{Papert91} by further
-emphasizing creation and tangible interaction using a tablet computer as a way
-to maintain the child's connection to their work (Fig.~\ref{fig:xo3}).
+%MSTONE PARALLEL STRUCTURE:
+%Children often learn from direct interactive experience. Consequently, Nell
+%intensifies our commitment to constructionism~\cite{Papert91} by further
+%emphasizing creation and tangible interaction using a tablet computer as a way
+%to maintain the child's connection to their work (Fig.~\ref{fig:xo3}).
+
+The constructionist learning philosophy emphasizes creation and
+tangible interaction.  Nell uses direct interaction on a tablet
+computer to maintain the child's connection to their work
+(Fig.~\ref{fig:xo3}).
 
@@
 ``TurtleScript'') are novel.
-
-\subsection{Handwriting recognition}
-To support \textit{direct interaction}, we use a separated-character
-handwriting recognition engine based on Hidden Markov Models~\cite{hanshu:handwriting}.
 
@@
 
+% MSTONE TEXT
+% Nell takes advantage of this fact 
+% to augment prior affordances for \textit{action}, \textit{reflection}, and
+% \textit{communication} with new affordances for \textit{engagement} in the form of
+% a modular \textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}.
+
 Automatic narrative generation has seen attention in

### 2012-03-11 paper.tex
Make 'new functionality' in Rahel's story more concrete, as mstone suggested.

@@
 \title{Growing up with Nell: A Narrative Interface for Literacy}
-%\subtitle{Narrative Interfaces for Literacy}
 %
@@
 the turtle is also used to write all the software running inside
-her special book.  Rahel begins to write stories with different
-kinds of new music functionality, which make the green book even more
-exciting for her younger brother and the other younger kids just now
-receiving them.
+her special book.  Rahel uses the blocks to write a new sort of rhythm
+game.  Her younger brother has just received his own green book,
+and Rahel writes him a story using her rhythm game to help him learn
+to count.
 

### 2012-03-11 paper.tex
Streamline end of 'Key Ideas' intro.
Lead with 'Children ____' for Narrative section, as m_stone proposed.

@@
 and is \textbf{Personalized} for, the child.
-Together, these four key concepts provide a novel
-experience which we use to address several challenges faced by earlier
-learning environments for children.
+Nell uses these four key concepts to build a novel learning platform
+which addresses several challenges we've encountered in earlier work.
 
@@
 
-Experience with our previous educational software
-system has shown gratifying uptake and
-use by children.  But we are often disappointed that some of our
-favorite pedagogical tools included in the systems were not being
-widely used.  Enthusiastic teachers would spark intensive use in
-certain classrooms, but children had difficulty finding their way into
-the material without strong guidance.
-
-Children (like all humans) are hard-wired for
-stories~\cite{boyd:stories}.  Nell uses a
+Children (like all humans) are hard-wired for stories~\cite{boyd:stories}.
+In our earlier learning software, we've seen some of our favorite
+pedagogical activities neglected because children had difficulty
+finding their way into the material without an enthusiastic teacher's
+guidance.  In response, Nell uses a
 \textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}
-to guide the child user through the pedagogical capabilities of the
-system. 
+to pave a path for the child user through the pedagogical content.
 In this interface, all system actions are shaped by a storybook metaphor, and
@@
 Each of the several characters represents a specific skill or subject
-area, and each adventure represents about a year's curriculum.  This
-design also improves modularity and allows updates and decentralized
+area, and each adventure represents about a year's curriculum.
+Adventures are further subdivided into story modules, which match the
+scope of a traditional lesson plan.  The serialized multi-character
+design improves modularity and allows updates and decentralized
 authorship.
@@
 active guidance (Apple-Guide--style~\cite{powers:appleguide}
-contextual help), or system services (switch activities; jump into a
+contextual help), or system services (switching activities; jumping into a
 related story module/lesson plan).

### 2012-03-11 paper.tex
Rahel's story uses the rhythm game, not Rahel.  Could be improved.
Temporarily comment out figure 2.

@@
 game.  Her younger brother has just received his own green book,
-and Rahel writes him a story using her rhythm game to help him learn
+and Rahel writes him a story which uses her rhythm game to help him learn
 to count.
@@
 
-\begin{figure}
-\centering
-\epsfig{file=roger1.eps, height=1.5in} % width=3.2in
-\caption{Diagram of Nell's primary interface.}\label{fig:nell}
-\end{figure}
+%\begin{figure}
+%\centering
+%\epsfig{file=roger1.eps, height=1.5in} % width=3.2in
+%\caption{Diagram of Nell's primary interface.}\label{fig:nell}
+%\end{figure}
 
@@
 % section 1.)  Move color changing into Related Work, even?
+
+%Children learn in different ways...
 

### 2012-03-11 paper.tex
Whoops, if we cut figure 2 we also need to cut the text referencing it.

@@
 
-\title{Growing up with Nell: A Narrative Interface for Literacy}
+\title{Growing Up with Nell: A Narrative Interface for Literacy}
 %
@@
 
-As depicted in Figure~\ref{fig:nell}, Nell's characters are
+%As depicted in Figure~\ref{fig:nell},
+Nell's characters are
 always-available agents layered above a particular system \textit{activity}

### 2012-03-12 paper.tex
use "Chris J. Ball" for cjb
Hm, I like Internet a bit better than Web.  Call me old-fashioned?
For context: the original idea was that a javascript on a web page would
be able to add a set of new production rules into your local Avatar Services,
since you can't evaluate the preconditions for a new rule set before you
know what they are.  But the distinction between "by visiting a website
some javascript runs to augment your story model" and "downloading from the
Web" is pretty probably not worth being too anal about.  So "incorporated"
may have been a little more accurate in my original draft,

@@
 \alignauthor
-C. Scott Ananian \qquad Chris Ball \qquad Michael Stone\titlenote{%
+C. Scott Ananian \qquad Chris J. Ball \qquad Michael Stone\titlenote{%
 Now at Akamai Technologies.}\\
@@
 are evaluated continuously to determine the actions of the narrative
-agents.  This production rules system
+agents.  This evaluation
 is made computationally efficient using the Rete
 algorithm.  Dialog is expressed using Curveship-style
-templates~\cite{montfort:curveship}.
+templates, allowing Nell to restructure
+expository tense, style, and point-of-view~\cite{montfort:curveship}.
 
@@
 
-New story modules can be downloaded from the Web in
+New story modules can be downloaded from the Internet in
 connected deployments.  In disconnected deployments, new stories

### 2012-03-12 paper.tex
Smooth Technology intro a bit.

@@
 \section{Technology}
-We describe the novel synthesis of technologies that makes up our current Nell
-prototype. Several of these technologies are well-established; a few (like
-``TurtleScript'') are novel.
+Our implementation choices further our goals of modularity and
+decentralization.  Nell's implementation is open source and uses
+standard web technologies to broaden the contributor base.  In this
+section we describe the technologies that enable Nell's four key ideas.
 
@@
 \end{figure}
-To support Nell's \textit{modular narrative interface}, we use a rule-based
-story model with programmable conflict resolution. (Fig.~\ref{fig:rules} shows
+To implement Nell's modular \textit{narrative interface}, we use a
+production rule system with programmable conflict resolution.
+Figure~\ref{fig:rules} shows
 an excerpt of the story module underlying the interaction described in
-Section~\ref{sec:lesson}.)
+Section~\ref{sec:lesson}.
 

### 2012-03-12 paper.tex
Rewrite the conflict-resolution section.

@@
 This makes it easy to take an existing story and modify it to
-better suit particular interests or a particular type of learner, or
+better suit particular interests, a particular type of learner---or
 just to add variety.  This can lead to conflicts: multiple versions of
-a story, or variant scenes within a story, may have their
+a story, or scenes within a story, may have their
 preconditions satisfied simultaneously.
@@
 invoked when a conflict exists; the actions associated with that scene
-will temporarily alter the priorities of the conflicting scenes to
-resolve the conflict.  For example, the \textit{story-intro-group}
-used in Fig.~\ref{fig:rules} defaults to a lower priority, such that
-continuing a story in progress is preferred to beginning a new scene.
-The resolution scene for \textit{story-intro-group} may be part of its
-own story: for example, it may bring us to the controls of a magic airship
-which allows us to pilot between locations corresponding to the
-various story continuations.  Selecting a destination location bumps
-the priority of the given scene allowing us to transition to it
-without further conflict.
+will (eventually) alter the priorities of the conflicting scenes to
+resolve the conflict.  The resolution scene may include any number of
+actions.  It may select randomly, or based a selection on
+child preferences, curriculum progress, or inferred learning style.
+The resolution scene may even initiate a new multi-scene story.
 
-Other rule groups may invoke random selection or select between stories
-based on child preferences, curriculum, or inferred learning style.
+For example, the \textit{story-intro-group}
+used in Figure~\ref{fig:rules} defaults to a low priority so that
+continuing a story in progress is preferred over starting a new story.
+When the story in progress is complete, several new story openers will be in
+conflict.  The resolution scene for \textit{story-intro-group} brings us
+to the controls of a magic airship which flies between locations
+corresponding to the various story continuations.  Eventually the
+child decides to fly to Mathland to hear the next episode in her
+agent Mathis' adventure, and that episode's priority is bumped to
+allow us to transition to it without further conflict.
 
-Nell's integrated story editor has not yet been implemented.  Its
-design is patterned after the Wide Ruled authoring tool, which was
-successfully used by non-technical story creators~\cite{skorupski:2009}.
+We plan to write an integrated story editor for Nell.  Our model will
+be the Wide Ruled authoring tool, which was successfully used by
+non-technical story creators~\cite{skorupski:2009}.
 

### 2012-03-12 paper.tex
s/code/core/.  (thanks kevix)

@@
 focuses on children even further from educational infrastructure.
-The narrative at Nell's code attempts to provide students additional
+The narrative at Nell's core attempts to provide students additional
 pedagogical guidance, and the direct interaction model avoids keyboard

### 2012-03-12 paper.tex
Massage the 'Extensible dialog' section.

@@
 \subsection{Extensible dialog}
-The child can directly interact with the Princess and other agents within
-Nell.  Convincing discourse is a hard problem, but
+Nell's use as an interactive diary and
+Portfolio~\cite{stefanakis:portfolios} is enhanced by the user's
+collaboration in the fiction that Nell is intelligent---what
+Turkle~\cite{turkle:alone} calls the ``ELIZA effect.''
+Convincing discourse is a hard problem, but
 reasonable approximations do not have to be difficult; even
@@
 % hours discussing her boyfriend.
-Nell's use as an interactive diary is enhanced by the child's
-collaboration in the fiction that Nell is intelligent---what
-Turkle~\cite{turkle:alone} calls the ``ELIZA effect.''
+We've chosen to allow the child to directly converse with the
+Princess and other agents within Nell.
 
@@
 
-We leverage AIML~\cite{aiml:2005},
-the markup language developed for the three-time Loebner Prize-winning
+Our implementation extends AIML~\cite{aiml:2005},
+the markup language developed for the three-time Loebner Prize--winning
 AliceBot.%
@@
   replacement~\cite{wilcox:2010}.  AIML's community and
-  diversity of third-party implementations make it preferable at present.}
+  diversity of third-party implementations make it our choice at present.}
 Each story module can include AIML fragments extending the
@@
 
-Fig.~\ref{fig:aiml}
-shows how this can be used to implement a traditional
-text adventure interaction for a particular story module.
-In this example the child can write instructions such as, ``go north,'' ``go
-east,'' or ``open door'' for the agent.  These trigger events
-handled by the story module to move the agent through the
-adventure---for example, by causing the activity to draw a
-new location, causing the agent to describe it, and shifting the
-AIML topic to enable vocabulary suitable for the new location.
+Figure~\ref{fig:aiml}
+shows an AIML fragment augmenting an agent with
+commands suitable for a text adventure story module.
+In such a story the child might write instructions such as
+\texttt{go north}, \texttt{go east}, or \texttt{open door} in
+addition to usual agent dialog.
+These new commands fire events which are referenced by scene
+preconditions.  For example, \texttt{north} may
+trigger a scene which causes the activity to draw a new location on
+the page, causes the agent to write and speak a description of the new
+location, and which shifts the AIML topic to enable additional vocabulary
+suitable for the new location.
 

### 2012-03-12 paper.tex
TurtleScript rewrite.

@@
 
+%%% XXX INTEGRATE THIS SECTION XXX
+Nell is implemented in standard Web technologies, including HTML5,
+WebGL, and JavaScript, with extensive use of offline caching.
+Resources are named by URL, even when disconnected from the
+internet.  This allows third parties to manage their own namespaces
+when extending Nell, and simplifies updates to story modules and
+the Nell system itself.
+%As far as possible, Nell is implemented in JavaScript.  
+% if the kid's going to learn some software stack, this seem like a
+% fine one to learn
+
 % it has a ``low floor and no ceiling'' to allow a child to master more
@@
 \subsection{TurtleScript}\label{sec:turtles}
-Nell's implementation leverages Web technologies, including HTML5,
-WebGL, and JavaScript, with extensive use of offline caching.
-Resources are named by URL, even when disconnected from the
-internet.  This allows third parties to manage their own namespaces
-when extending Nell, and simplifies updates to story modules and
-the Nell system itself.
-% if the kid's going to learn some software stack, this seem like a
-% fine one to learn
 
-As far as possible, Nell is implemented in JavaScript.  We have
-developed a block-oriented view of JavaScript, which we call
-TurtleScript~\cite{turtlescript}.  Combined with subsets of JavaScript
-for beginners, this allows us to use a single implementation language
-which scales from logo-like tasks all the way through the
-implementation of the entire Nell system itself.  TurtleScript's block
-view of JavaScript is isomorphic to the standard textual
-representation of code.  We have begun implementation of a full
-programming environment for TurtleScript which is naturally-suited to
-tablet machines.
+We have developed a block-oriented view of JavaScript, which we call
+TurtleScript (Figure~\ref{fig:turtlescript}).  Combined with subsets of JavaScript
+for learners, this allows us to use TurtleScript as the single
+implementation language for Nell.  TurtleScript scales from Logo-like
+pedagogical tasks up to the implementation of TurtleScript and the
+Nell system itself.  TurtleScript's block view is isomorphic to
+JavaScript's standard textual
+representation, which preserves the code-reading skills of existing
+programmers.  We have begun implementation of a
+tablet programming environment for TurtleScript which uses
+drag-and-drop, direct interaction, and intelligent prompting to reduce
+text entry requirements.
 
 % XXX refer to TurtleScript excerpt in figure somewhere here.
+\begin{figure}
+\centering
+\epsfig{file=turtlescript.eps, height=1.5in}
+\caption{Factorial function in TurtleScript~\cite{turtlescript}.}\label{fig:turtlescript}
+\end{figure}
 
-Our goal is to allow the child to eventually dig as far into Nell's
-implementation as their capabilities allow.  Although the operating system and
-the browser itself are inaccessible from the web platform, we can
-provide full access above that level.  The existence of meta-circular
-interpreters for JavaScript~\cite{narcissus} implies that we can
-expose the system all the way down to the programming language
-implementation.  Via TurtleScript, this code is all visible in a
-friendly block-based form.
+TurtleScript allows the child to dig as far into Nell's implementation
+as their capabilities allow.  Although the operating system and the
+browser implementation remain inaccessible, incorporating a
+meta-circular interpreter for JavaScript (such as \cite{narcissus}) allows us
+to expose the system all the way down to the programming
+language implementation,
+% Via TurtleScript, this code is all visible in a
+% friendly block-based form.
 

### 2012-03-12 paper.tex
Tweaks to TurtleScript section.

@@
 We have developed a block-oriented view of JavaScript, which we call
-TurtleScript (Figure~\ref{fig:turtlescript}).  Combined with subsets of JavaScript
-for learners, this allows us to use TurtleScript as the single
+TurtleScript~\cite{turtlescript}.  This allows us to use
+the TurtleScript view of JavaScript as the single
 implementation language for Nell.  TurtleScript scales from Logo-like
-pedagogical tasks up to the implementation of TurtleScript and the
+pedagogical tasks (in subsets of the language) up to the
+implementation of TurtleScript and the
 Nell system itself.  TurtleScript's block view is isomorphic to
 JavaScript's standard textual
-representation, which preserves the code-reading skills of existing
-programmers.  We have begun implementation of a
+representation to preserve the code-reading skills of existing
+programmers (Figure~\ref{fig:turtlescript}).  We have begun implementation of a
 tablet programming environment for TurtleScript which uses
@@
 \epsfig{file=turtlescript.eps, height=1.5in}
-\caption{Factorial function in TurtleScript~\cite{turtlescript}.}\label{fig:turtlescript}
+\caption{Factorial function in TurtleScript.}\label{fig:turtlescript}
 \end{figure}

### 2012-03-12 paper.tex
Photoshop Nell into the XO-3 photo.
Tweak conclusion.

@@
 \centering
-\epsfig{file=xo3-3.eps, width=3.2in}
+\epsfig{file=xo3-4.eps, width=3.2in}
 \caption{Nell's solar-powered hardware platform.  The tablet design
@@
 its child owner.  Personalized direct interface engages the child, and
-narrative guides them through Nell's pedagogic material.  The low-cost
+narrative guides them through the pedagogic material.  The low-cost
 solar-powered hardware allows Nell to reach further into the
@@
 We are in the process of implementing the Nell system, and expect to
-do initial deployments in Africa later this year.
+field initial deployments of subsets of Nell in Africa later this year.
 

### 2012-03-12 paper.tex
Fix a bad LaTeX line break.

@@
 contextual help), or system services (switching activities; jumping into a
-related story module/lesson plan).
+related story module).
 

### 2012-03-12 paper.tex
Unabbreviate "Figure" -- we've got the space.
Turn off authors for submission, so that IDC can do their blind reviewing thing.
Copy-editing the text at 4:30am.  Hope I'm getting everything.

@@
 \usepackage{url}
-\usepackage{ifthen}\newboolean{noauthors}\setboolean{noauthors}{false}
+\usepackage{ifthen}\newboolean{noauthors}\setboolean{noauthors}{true}
 
@@
 teaching about itself, and a capable child can dive into Nell's source
-code and make meaningful changes.  In section~\ref{sec:turtles} we
-describe how this goal is approached.
+code and make meaningful changes.
+%In Section~\ref{sec:turtles} we describe how this goal is approached.
 
@@
 computer to maintain the child's connection to their work
-(Fig.~\ref{fig:xo3}).
+(Figure~\ref{fig:xo3}).  In particular, Nell uses handwriting
+recognition as a primary interface.
 
@@
 to learn an abstract touchpad or mouse interface.  It also supports
-our literacy goals by allowing direct handwriting instruction and the
-use of handwriting as an interface.
+our literacy goals by allowing direct handwriting instruction and
+development of motor skills.
 
 \section{Technology}
-Our implementation choices further our goals of modularity and
-decentralization.  Nell's implementation is open source and uses
+Our technology choices further our goals of modularity and
+decentralization.  Nell's implementation is free and open source and uses
 standard web technologies to broaden the contributor base.  In this
@@
 
-Story modules can be inherited and extended, as shown in the figure.
+Story modules can be inherited and extended.
 This makes it easy to take an existing story and modify it to
@@
 resolve the conflict.  The resolution scene may include any number of
-actions.  It may select randomly, or based a selection on
+actions.  It may select randomly among the conflicting scenes or base
+its selection on
 child preferences, curriculum progress, or inferred learning style.
@@
 continuing a story in progress is preferred over starting a new story.
-When the story in progress is complete, several new story openers will be in
+When a story module is completed, several new story openings will be in
 conflict.  The resolution scene for \textit{story-intro-group} brings us
 to the controls of a magic airship which flies between locations
-corresponding to the various story continuations.  Eventually the
-child decides to fly to Mathland to hear the next episode in her
-agent Mathis' adventure, and that episode's priority is bumped to
-allow us to transition to it without further conflict.
+corresponding to the various story continuations.  In that set of
+scenes, the child may eventually decide to fly to Mathland to hear the
+next episode in her agent Mathis' adventure, which would bump up the
+appropriate episode's priority to allow us to make the transition
+without further conflict.
 
-We plan to write an integrated story editor for Nell.  Our model will
-be the Wide Ruled authoring tool, which was successfully used by
+Nell will also include an integrated story editor.  Our exemplar is
+the Wide Ruled authoring tool, which was successfully used by
 non-technical story creators~\cite{skorupski:2009}.
@@
 Our implementation extends AIML~\cite{aiml:2005},
-the markup language developed for the three-time Loebner Prize--winning
+the markup language developed for the Loebner Prize--winning
 AliceBot.%
@@
 \texttt{go north}, \texttt{go east}, or \texttt{open door} in
-addition to usual agent dialog.
+addition to the usual dialog with the agent.
 These new commands fire events which are referenced by scene
@@
 the page, causes the agent to write and speak a description of the new
-location, and which shifts the AIML topic to enable additional vocabulary
-suitable for the new location.
+location, and which causes a shift in the AIML \textit{topic} to
+enable additional vocabulary suitable for the new location.
 
@@
 We have developed a block-oriented view of JavaScript, which we call
-TurtleScript~\cite{turtlescript}.  This allows us to use
-the TurtleScript view of JavaScript as the single
+TurtleScript~\cite{turtlescript}.
+The TurtleScript view of JavaScript is the single
 implementation language for Nell.  TurtleScript scales from Logo-like
@@
 implementation of TurtleScript and the
-Nell system itself.  TurtleScript's block view is isomorphic to
+Nell system itself.  As Figure~\ref{fig:turtlescript} shows,
+TurtleScript's block view is isomorphic to
 JavaScript's standard textual
 representation to preserve the code-reading skills of existing
-programmers (Figure~\ref{fig:turtlescript}).  We have begun implementation of a
+programmers.  We have begun implementation of a
 tablet programming environment for TurtleScript which uses
@@
 to expose the system all the way down to the programming
-language implementation,
+language implementation.
 % Via TurtleScript, this code is all visible in a
@@
 The TinkRBook system~\cite{chang:tinkrbook} shares a similar focus on
-personalization and literacy.  Nell attempts to scale these
+personalization and literacy.  Nell attempts to scale up these
 ideas with a modular extensible narrative system and a broader range
@@
 The Sugar Learning Platform~\cite{sugar} is an earlier effort at
-child education.  We've learned directly from this effort.  Nell
-focuses on children even further from educational infrastructure.
+child education which we've been directly involved in.  Nell's focus is
+on children even further from educational infrastructure than those
+targeted by Sugar.
 The narrative at Nell's core attempts to provide students additional
@@
 Automatic narrative generation has seen attention in
-gaming, simulation, and training contexts~\cite{riedl:2008}.  Nell's
-use of these ideas in a pedagogical context is unique.
+gaming, simulation, and training contexts~\cite{riedl:2008}.
+We believe
+Nell's use of these ideas in a pedagogical context is unique.
 
@@
 narrative guides them through the pedagogic material.  The low-cost
-solar-powered hardware allows Nell to reach further into the
-least-developed areas of the world, to help those far from traditional
+solar-powered deployment hardware allows Nell to reach further into the
+least-developed areas of the world to help those without traditional
 educational infrastructure.

### 2012-03-12 paper.tex
Spell-checking.
Rewrite Personalization section to lead with the Multiple Intelligences stuff.

@@
 contextual help), or system services (switching activities; jumping into a
-related story module).
+related story module).  A system-wide \textit{achievement} system provides
+goals and rewards.
 
@@
 
-%Children learn in different ways...
+Children learn in different ways.  Nell provides multiple alternative
+story modules for its plot points/lessons to engage the child's
+interests and cater to multiple intelligences~\cite{gardner:mi}.
+Selecting between the alternatives can depend on either explicit
+choices made by the child (``fractions in outer space,'' if the child
+chose a space-themed story or previously indicated an interest in space)
+or prior success with a given
+lesson style (a large number of accomplishments in musical-rhythmic
+tasks suggests taking a rhythmic approach to fractions).  Similarly, several
+different achievements can coexist, rewarding different ways of
+accomplishing the same pedagogical goal.
 
-Children often decorate their belongings, and the
-child owning a Nell system is encouraged to personalize it.
-Superficially, colors can be changed and characters can be renamed and
-dressed up.  But Nell also supports deeper story personalization.
+This record of past choices and accomplishments is stored in a
+\textit{Journal} and can be reviewed by the child.  The contents of the
+Journal can be rearranged to create a Portfolio~\cite{stefanakis:portfolios}
+demonstrating the child's progress.
+In order to encourage fearless play and experimentation, the Journal
+also supports pervasive undo; the child can rewind and replay the
+narrative starting at any past point in the journal.
 
-\textit{Avatar Services} stores all choices made by the child and make them
-accessible to story modules and activities.  Accomplishments
-(completed lessons) are also stored.  Nell can thus provide multiple
-alternative story modules for a given plot point/lesson, and select
-between them based on either the explicit choices made by the child
-(``fractions in outer space,'' if the child previously indicated a
-preference for space-themed lessons) or prior success with a given
-lesson style (a large number of accomplishments in musical-rhythmic
-tasks suggests a rhythmic approach to fractions).  In this manner we
-can engage the child's interests and cater to different learning
-styles.%~\cite{gardner:mi}.%%%%%%% <- lead with this.
-
-Moving further, this record of past choices and accomplishments can be
-reviewed.  The narrative history is stored in the \textit{Journal},
-where it can form the core of a Portfolio~\cite{stefanakis:portfolios}.
-In order to encourage fearless play and experimentation, the Journal
-also supports pervasive undo; you can replay the narrative starting at
-any past point in the journal.
-
-In the future, we expect to allow further narrative personalization
-via recombination of story elements.
+Content can also be remixed for further personalization.
+The child is encouraged to rename characters and change clothing,
+colors, and other superficial details.  In the future
+we expect to allow further narrative customization by
+recombination of story elements.
 Lebowitz~\cite{lebowitz:universe85} and Riedl~\cite{riedl:planning}
@@
 attempting to generate thousands of stories from tens of templates, we
-expect to select and then modestly adapt from hundreds of story
+hope to select and then modestly adapt from hundreds of story
 modules created in a decentralized manner by teachers---and eventually
@@
 
+% CSA: three axes of growth
+%  1. serial tech tree of story material; more advanced topics
+%  2. turtlescript -- metacircular understanding of the system itself
+%  3. community -- no limit to content creation, including by the child.
+
 Children grow. To accommodate them, Nell provides a ``low floor and
-no ceiling.''  The low
-floor is a very simple base system, usable with no literacy skills or
+no ceiling.''  A very simple base system provides the low
+floor, usable with no literacy skills or
 external guidance.  The serial progression of story modules grows with
@@
 
-%%  LocalWords:  personalization Lebowitz
+%%  LocalWords:  personalization Lebowitz TurtleScript AIML Loebner
+%%  LocalWords:  AliceBot modularity Rahel Ananian Akamai Papert JSON
+%%  LocalWords:  Riedl WebGL touchpad Curveship Rete TinkRBook

### 2012-03-12 paper.tex
Move Narcissus into a footnote to bring Bibliography back down to 1 column.

@@
 browser implementation remain inaccessible, incorporating a
-meta-circular interpreter for JavaScript (such as \cite{narcissus}) allows us
-to expose the system all the way down to the programming
+meta-circular interpreter for JavaScript%
+\footnote{For example, \url{https://github.com/mozilla/narcissus/}}
+% (such as \cite{narcissus})
+allows us to expose the system all the way down to the programming
 language implementation.

### 2012-03-12 paper.tex
Rewrite "Growable" section.

@@
 
-Children grow. To accommodate them, Nell provides a ``low floor and
-no ceiling.''  A very simple base system provides the low
-floor, usable with no literacy skills or
-external guidance.  The serial progression of story modules grows with
-the child to tackle topics of increasing difficulty.
-Later, the ability for third
-parties to independently author story modules ensures that
-Nell can continue to grow.  Finally, there are no external dependencies
-limiting a child's ability to participate in this process: a Nell
-system contains a story editor and everything necessary to
+Children grow.  Nell aims to provide a ``low floor and no ceiling'' to
+grow with them, along three main axes.
+%There are three main axes of growth.
+%  Nell's low floor is its
+%simple base system,
+%usable with no literacy skills or external guidance.
+%From there Nell grows in three directions.
+
+First, Nell tracks the child's growing intelligence and capability
+with increasingly challenging story material.  As the child
+accumulates achievements and demonstrates
+proficiency, the serial story moves on to more advanced topics.
+
+Second, Nell seeks to grow its authoring community.  Decentralized
+authorship ensures that Nell's instructional content
+will continue to grow, with an increasing number of topics and
+customizations.  There are no
+ceilings barring a child from participating: Nell
+contains a story editor and everything necessary to
 author and publish story modules.
 
-We also seek to eliminate the ``ceiling'' between the children and
-the authors of the Nell system itself.  Nell should be capable of
-teaching about itself, and a capable child can dive into Nell's source
-code and make meaningful changes.
+To further promote collaboration
+Nell is implemented in standard Web technologies (HTML5,
+WebGL, and JavaScript) with offline caching.
+Resources are named by URL, even when disconnected from the
+internet, which simplifies the distribution of updates to story modules and
+the Nell system.  URL-based naming allows third parties
+to manage their own namespaces when extending Nell.
+
+Finally, Nell is designed to allow a capable child to learn
+the construction of the system itself,
+eliminating the ceiling between the child and
+the authors of the Nell system.  Nell can teach
+about itself, and its source code is open for exploration within Nell.
+Meaningful changes can be performed without external tools.
+
+%.  Meaningful changes to the Nell system can
+%be performed on the system itself.
 %In Section~\ref{sec:turtles} we describe how this goal is approached.
 
-%%% XXX INTEGRATE THIS SECTION XXX
-Nell is implemented in standard Web technologies, including HTML5,
-WebGL, and JavaScript, with extensive use of offline caching.
-Resources are named by URL, even when disconnected from the
-internet.  This allows third parties to manage their own namespaces
-when extending Nell, and simplifies updates to story modules and
-the Nell system itself.
 %As far as possible, Nell is implemented in JavaScript.

### 2012-03-12 paper.tex
Final copy editing, tweaking line breaks, removing overful hboxen, etc.

@@
 \usepackage{ifthen}\newboolean{noauthors}\setboolean{noauthors}{true}
+\hyphenation{Web-GL Java-Script Turtle-Script infra-structure}
 
@@
 She nods and listens to a story about a princess.  Later, when
-the Rahel has learned a little more, she will tell the machine that the
+the girl has learned a little more, she will tell the machine that the
 princess is named ``Rahel'' like she is and that she likes to wear blue---but
@@
 \textit{Narrative Interface}~\cite{bizzocchi:narrative,don:narrative}
-to pave a path for the child user through the pedagogical content.
-In this interface, all system actions are shaped by a storybook metaphor, and
-interactions with the system are conceptualized as interactions with one of the
+to pave a path for the child user through the learning material.
+In the narrative interface, all system actions are shaped by a
+storybook metaphor, and
+interactions with the system are framed as interactions with one of the
 system protagonists.
@@
 
-Children learn in different ways.  Nell provides multiple alternative
+Children learn in different ways.  Nell provides multiple
 story modules for its plot points/lessons to engage the child's
@@
 lesson style (a large number of accomplishments in musical-rhythmic
-tasks suggests taking a rhythmic approach to fractions).  Similarly, several
+tasks suggests a rhythmic approach to fractions).  Similarly, several
 different achievements can coexist, rewarding different ways of
@@
 
-This record of past choices and accomplishments is stored in a
+The record of past choices and accomplishments is stored in a
 \textit{Journal} and can be reviewed by the child.  The contents of the
@@
 
-Content can also be remixed for further personalization.
+Content can be remixed for further personalization.
 The child is encouraged to rename characters and change clothing,
 colors, and other superficial details.  In the future
-we expect to allow further narrative customization by
+we expect to accomplish further narrative customization by
 recombination of story elements.
@@
 authorship ensures that Nell's instructional content
-will continue to grow, with an increasing number of topics and
+continues to increase its topics and
 customizations.  There are no
-ceilings barring a child from participating: Nell
+ceilings between children and teachers: Nell
 contains a story editor and everything necessary to
@@
 
-To further promote collaboration
-Nell is implemented in standard Web technologies (HTML5,
-WebGL, and JavaScript) with offline caching.
+To further promote collaboration,
+Nell is free and open source and implemented in standard web
+technologies (WebGL, HTML5, and JavaScript) with offline caching.
 Resources are named by URL, even when disconnected from the
 internet, which simplifies the distribution of updates to story modules and
-the Nell system.  URL-based naming allows third parties
+the Nell system.  URL-based identifiers also allow third parties
 to manage their own namespaces when extending Nell.
 
-Finally, Nell is designed to allow a capable child to learn
+Finally, Nell is designed to allow a capable child to learn about
 the construction of the system itself,
@@
 \section{Technology}
-Our technology choices further our goals of modularity and
-decentralization.  Nell's implementation is free and open source and uses
-standard web technologies to broaden the contributor base.  In this
+%Our technology choices further our goals of modularity and
+%decentralization.  Nell's implementation is free and open source and uses
+%standard web technologies to broaden the contributor base.  In this
+%section we describe the technologies that enable Nell's four key ideas.
+
+Our implementation choices further our goals of modularity and
+decentralization.  In this
 section we describe the technologies that enable Nell's four key ideas.
@@
 \end{figure}
-To implement Nell's modular \textit{narrative interface}, we use a
+To implement Nell's modular narrative interface, we use a
 production rule system with programmable conflict resolution.
@@
 actions.  It may select randomly among the conflicting scenes or base
-its selection on
-child preferences, curriculum progress, or inferred learning style.
+its selection on the
+child's preferences, curriculum progress, or inferred learning style.
 The resolution scene may even initiate a new multi-scene story.
@@
 used in Figure~\ref{fig:rules} defaults to a low priority so that
-continuing a story in progress is preferred over starting a new story.
+continuing a story in progress is preferred above starting a new story.
 When a story module is completed, several new story openings will be in
-conflict.  The resolution scene for \textit{story-intro-group} brings us
-to the controls of a magic airship which flies between locations
+conflict.  The resolution scene for \textit{story-intro-group} may bring us
+to the controls of an airship which flies between locations
 corresponding to the various story continuations.  In that set of
@@
 next episode in her agent Mathis' adventure, which would bump up the
-appropriate episode's priority to allow us to make the transition
+appropriate scene's priority to allow us to make the transition
 without further conflict.
 
-Nell will also include an integrated story editor.  Our exemplar is
+Nell will include an integrated story editor.  Our exemplar is
 the Wide Ruled authoring tool, which was successfully used by
@@
 Nell's use as an interactive diary and
-Portfolio~\cite{stefanakis:portfolios} is enhanced by the user's
+\textit{portfolio}~\cite{stefanakis:portfolios} is enhanced by the child's
 collaboration in the fiction that Nell is intelligent---what
@@
 the page, causes the agent to write and speak a description of the new
-location, and which causes a shift in the AIML \textit{topic} to
+location, and finally causes a shift in the AIML \textit{topic} to
 enable additional vocabulary suitable for the new location.
@@
 TurtleScript~\cite{turtlescript}.
-The TurtleScript view of JavaScript is the single
+The TurtleScript dialect of JavaScript is the single
 implementation language for Nell.  TurtleScript scales from Logo-like
@@
 The TinkRBook system~\cite{chang:tinkrbook} shares a similar focus on
-personalization and literacy.  Nell attempts to scale up these
-ideas with a modular extensible narrative system and a broader range
+personalization and literacy.  Nell attempts to scale the system up
+with a modular extensible narrative system and a broader range
 of pedagogic material.
@@
 The Sugar Learning Platform~\cite{sugar} is an earlier effort at
-child education which we've been directly involved in.  Nell's focus is
-on children even further from educational infrastructure than those
+child education in which we've been directly involved.  Nell's focus is
+on children further from educational infrastructure than those
 targeted by Sugar.
-The narrative at Nell's core attempts to provide students additional
-pedagogical guidance, and the direct interaction model avoids keyboard
+The narrative at Nell's core attempts to provide students
+pedagogical guidance missing from Sugar,
+and the direct interaction model avoids keyboard
 problems which have plagued Sugar's initial deployments.  Nell also
@@
 We believe
-Nell's use of these ideas in a pedagogical context is unique.
+Nell's synthesis of these ideas in a pedagogical context is unique.
 
@@
 We have described the design of a novel narrative direct-interface
-system for teaching literacy, which is personalized for and grows with
+system for education, beginning with literacy, which is personalized for and grows with
 its child owner.  Personalized direct interface engages the child, and
-narrative guides them through the pedagogic material.  The low-cost
-solar-powered deployment hardware allows Nell to reach further into the
+narrative guides them through pedagogic material.  Low-cost
+solar-powered hardware allows Nell to reach further into the
 least-developed areas of the world to help those without traditional

### 2012-03-12 paper.tex
Overful hboxes DIE DIE DIE.

@@
 authorship ensures that Nell's instructional content
-continues to increase its topics and
+continues to increase in the number of topics and
 customizations.  There are no
@@
 Nell is free and open source and implemented in standard web
-technologies (WebGL, HTML5, and JavaScript) with offline caching.
+technologies (JavaScript, HTML5, and WebGL) with offline caching.
 Resources are named by URL, even when disconnected from the

### 2012-03-14 paper.tex
Turn authors back on, shrink TurtleScript to keep everything on 4 pages.
Attempt to make borders on TurtleScript diagram more clear.
Turn authors off, resubmit with darker outlines for TurtleScript.
Avoid unmotivated praise, which demotivates.
See http://www.alfiekohn.org/parenting/gj.htm (from Mitch)
and
http://www.amazon.com/Self-theories-Motivation-Personality-Development-Psychology/dp/1841690244/
(from Caroline).

@@
 Princess asks, \textit{``Can you find something Red?''} and Rahel uses the
-camera to photograph berries on a nearby bush.
+camera to photograph a berry on a nearby bush.
 
-\textit{``Very good!  You are a smart girl!''}
+\emph{``Good work!  I see a little red \underline{here}.  Can you find
+something \emph{big} and red?''}
+% avoid excessive/unmerited praise
+% try to say *what* you like, rather than just vague praise
+% provide quality feedback, encourage the kid to max out the meter on
+% their own (ie, making the kid proud of himself for doing well is
+% better than telling him they're doing well)
 
@@
 next episode in her agent Mathis' adventure, which would bump up the
-appropriate scene's priority to allow us to make the transition
+appropriate scene's priority to make the transition
 without further conflict.
@@
 \centering
-\epsfig{file=turtlescript.eps, height=1.5in}
+\epsfig{file=turtlescript-2.eps, height=1.3in}
 \caption{Factorial function in TurtleScript.}\label{fig:turtlescript}

### 2012-03-15 paper.tex
Ask for help?
Proposed a JSON format for AIML. (Blame Alan Kay.)

@@
 \begin{verbatim}
-<aiml version="1.0">
-<category>
- <pattern>GO *</pattern>
- <template><srai><star/></srai></template>
-</category>
-
-<category>
- <pattern>N</pattern>
- <template><srai>NORTH</srai></template>
-</category>
-
-<category>
- <pattern>NORTH</pattern>
- <template>
-  <javascript>FireEvent(Chat, "North")</javascript>
- </template>
-</category>
-</aiml>
+{ aiml: {
+  version: "1.0",
+  categories: [
+    { pattern: "GO *",
+      template: "{*}" },
+    { pattern: "N",
+      template: "{NORTH}" },
+    { pattern: "NORTH",
+      template: js`FireEvent(Chat, "NORTH")` },
+    { pattern: "* THE DOOR",
+      template: "{*} DOOR" },
+    { pattern: "* SQUEAKY DOOR",
+      template: "{*} DOOR" },
+    { pattern: "OPEN DOOR",
+      template: js`FireEvent(Chat, "OPEN DOOR")` }
+  ] } }
 \end{verbatim}
 \caption{An AIML fragment which recognizes the commands \texttt{N},
-  \texttt{North}, and \texttt{Go North}. The \texttt{<srai>} tag invokes the
-  recognizer recursively.}\label{fig:aiml}
+  \texttt{North}, \texttt{Go North}, \texttt{Open the squeaky door}
+  and variants.  AIML's XML format has been translated to JSON.
+  Template contents in braces invoke the recognizer recursively.
+% XXX: how to handle multiple stars?  In XML, it's <star index="1">
+}\label{fig:aiml}
 \end{figure}
@@
 field initial deployments of subsets of Nell in Africa later this year.
+Volunteer assistance is welcome.
 

### 2012-03-15 paper.tex
Describe the JSON translation of AIML.  It's Alan Kay's fault.

@@
 \end{verbatim}
-\caption{An AIML fragment which recognizes the commands \texttt{N},
+\caption{An AIML fragment which recognizes the commands
+  %\texttt{N},
   \texttt{North}, \texttt{Go North}, \texttt{Open the squeaky door}
-  and variants.  AIML's XML format has been translated to JSON.
-  Template contents in braces invoke the recognizer recursively.
+  and variants.  AIML's XML syntax has been translated to JSON.
+  Template contents in braces invoke the recognizer recursively
+  (what AIML calls \texttt{<srai>}).
 % XXX: how to handle multiple stars?  In XML, it's <star index="1">
@@
 conversational capabilities of Nell's agents.
+% we say this in the figure, don't repeat it here? ==>
+%    We have translated AIML's XML-based syntax to JSON.
 
@@
 The TurtleScript dialect of JavaScript is the single
-implementation language for Nell.  TurtleScript scales from Logo-like
+implementation language for Nell.%
+\footnote{Since JSON is the data-oriented subset of JavaScript,
+our examples in Figures~\ref{fig:rules} and \ref{fig:aiml} are valid
+TurtleScript.} % XXX: use TurtleScript tiles for the figures?
+TurtleScript scales from Logo-like
 pedagogical tasks (in subsets of the language) up to the

### 2012-03-15 paper.tex
Rephrase the reason for JavaScript isomorphism.  (More Alan Kay feedback.)

@@
 JavaScript's standard textual
-representation to preserve the code-reading skills of existing
-programmers.  We have begun implementation of a
+representation to
+% CSA: hard to state all the reasons in just one short phrase, sigh.
+ensure the readability of and compatibility with existing JavaScript libraries.
+%preserve the code-reading skills of existing programmers.
+%
+We have begun implementation of a
 tablet programming environment for TurtleScript which uses

### 2012-04-23 paper.tex
Alternate non-quasi-quote template syntax.
This uses ${} so that it's compatible with quasi-quote in the future.
ie, eventually "${foo}" might become `${foo}`.
Formatting tweaks for final submission.
Quote keys in JSON (these are optional in JS, but mandatory in JSON).

@@
 \usepackage{url}
-\usepackage{ifthen}\newboolean{noauthors}\setboolean{noauthors}{true}
+\usepackage{ifthen}\newboolean{noauthors}\setboolean{noauthors}{false}
 \hyphenation{Web-GL Java-Script Turtle-Script infra-structure}
@@
 
-\title{Growing Up with Nell: A Narrative Interface for Literacy}
+\title{Growing Up With Nell: A Narrative Interface for Literacy}
 %
@@
 % number that will appear in the \additionalauthors section.
+
+% Copyright info
+\toappear{\textit{IDC 2012}, June 12--15, 2012, Bremen, Germany.\\
+Copyright 2012 ACM 978-1-4503-1007-9\ldots\$10.00.}
 
@@
 \begin{verbatim}
-{ world: { // id gives canonical URL for this module
-    id: "/nell.laptop.org/chapters/alphabet-book",
-    scene: "opening", // initial scene
-    inherit_from: "/nell.laptop.org/core/drawing_activity",
-  }, scenes: [
-    { id: "opening", subject: "actors/princess",
-      verb: "speak", object: "actors/user",
-      text: "Hello! Would (o/2) like to hear a story?",
-      group: "/nell.laptop.org/core/story-intro-group",
+{ "world": { // id gives canonical URL for this module
+    "id": "/nell.laptop.org/chapters/alphabet-book",
+    "scene": "opening", // initial scene
+    "inherit_from": "/nell.laptop.org/core/drawing_activity",
+  }, "scenes": [
+    { "id": "opening", subject: "actors/princess",
+      "verb": "speak", object: "actors/user",
+      "text": "Hello! Would (o/2) like to hear a story?",
+      "group": "/nell.laptop.org/core/story-intro-group",
     },
-    { id: "story-1",
-      cond: rule`At(opening) && Event(Chat, Yes)`,
-      subject: "actors/princess",
-      verb: "speak", object: "actors/user",
-      text: "Once upon a time..." },
+    { "id": "story-1",
+      "cond": rule`At(opening) && Event(Chat, Yes)`,
+      "subject": "actors/princess",
+      "verb": "speak", object: "actors/user",
+      "text": "Once upon a time..." },
     ...
-    { id: "r-drawn",
-      cond: rule`Event(Activity, DrawR)`,
-      subject: "actors/princess",
-      verb: "speak", object: "actors/user",
-      text: "Very good, (o/3)!",
-      action: js`startAnimation("r-runs")` }, ...
+    { "id": "r-drawn",
+      "cond": rule`Event(Activity, DrawR)`,
+      "subject": "actors/princess",
+      "verb": "speak", object: "actors/user",
+      "text": "Very good, (o/3)!",
+      "action": js`startAnimation("r-runs")` }, ...
   ], ... }
@@
 \begin{verbatim}
-{ aiml: {
-  version: "1.0",
-  categories: [
-    { pattern: "GO *",
-      template: "{*}" },
-    { pattern: "N",
-      template: "{NORTH}" },
-    { pattern: "NORTH",
-      template: js`FireEvent(Chat, "NORTH")` },
-    { pattern: "* THE DOOR",
-      template: "{*} DOOR" },
-    { pattern: "* SQUEAKY DOOR",
-      template: "{*} DOOR" },
-    { pattern: "OPEN DOOR",
-      template: js`FireEvent(Chat, "OPEN DOOR")` }
+{ "aiml": {
+  "version": "1.0",
+  "categories": [
+    { "pattern": "GO *",
+      "template": "{*}" },
+    { "pattern": "N",
+      "template": "{NORTH}" },
+    { "pattern": "NORTH",
+      "template": js`FireEvent(Chat, "NORTH")` },
+    { "pattern": "* THE DOOR",
+      "template": "{*} DOOR" },
+    { "pattern": "* SQUEAKY DOOR",
+      "template": "{*} DOOR" },
+    { "pattern": "OPEN DOOR",
+      "template": js`FireEvent(Chat, "OPEN DOOR")` }
   ] } }
 \end{verbatim}
+% alternate template syntax, compatible w/ quasiquote but not
+% dependent on it:
+% template: "${srai(star(0))}" // can be abbreviated as srai(star()) or srai()
+% template: "${srai("NORTH"}}"
+% template: "${FireEvent(Chat, "NORTH")}"
+% template: "${srai()} DOOR" // abbreviation for srai(star(0))
 \caption{An AIML fragment which recognizes the commands

### 2012-04-23 paper.tex
Space-saving hacks.
Remove Weizenbaum cite, since Turkle mentions this in the previous citation
and Weizenbaum doesn't really talk much about Eliza in the cited book.

Remove middle initial from E. H. Stefanakis in order to fit citation onto
two lines instead of three.

@@
 the rudimentary conversational abilities of ELIZA elicited hours of
-conversation~\cite{weizenbaum:power}.
+conversation.%~\cite{weizenbaum:power}
 % her secretary is divulging her life story
@@
 \begin{figure}\small
+% (could remove
+%  "version": "1.0",
+% before "categories" in order to save space in paper)
 \begin{verbatim}

### 2012-04-23 paper.tex
Further tweaks to reduce external dependencies and conform to final style.
Inline .bib file to create "self-contained" .tex, per ACM instructions.
The eps figures are not inlined, I guess that's okay.

@@
 %\documentclass{acm_proc_article-sp}
-\documentclass[preprint]{sig-alternate}
+\documentclass{sig-alternate}
 \usepackage{url}
-\usepackage{ifthen}\newboolean{noauthors}\setboolean{noauthors}{false}
 \hyphenation{Web-GL Java-Script Turtle-Script infra-structure}
 
-\begin{document}\toappear{}
+\begin{document}
 
@@
 
-\ifthenelse{\boolean{noauthors}}{}{
 \numberofauthors{1}
@@
 }
-}
 \date{12 March 2012}
@@
 
+\conferenceinfo{IDC 2012,}{June 12--15, 2012, Bremen, Germany}
+\CopyrightYear{2012}
+\crdata{978-1-4503-1007-9}
+
 % Copyright info
-\toappear{\textit{IDC 2012}, June 12--15, 2012, Bremen, Germany.\\
-Copyright 2012 ACM 978-1-4503-1007-9\ldots\$10.00.}
+%\toappear{\textit{IDC 2012}, June 12--15, 2012, Bremen, Germany.\\
+%Copyright 2012 ACM 978-1-4503-1007-9\ldots\$10.00.}
 
@@
 \bibliographystyle{abbrv}
-\bibliography{paper}  % paper.bib is the name of the Bibliography in this case
-% You must have a proper ".bib" file
-%  and remember to run:
-% latex bibtex latex latex
-% to resolve all references
+
+% ACM needs 'a single self-contained file'!
+%\bibliography{paper}  % paper.bib is the name of the Bibliography in this case
+\begin{thebibliography}{10}
+
+\bibitem{aiml:2005}
+{A.L.I.C.E. AI Foundation}.
+\newblock Artificial intelligence markup language ({AIML}) version 1.0.1.
+\newblock \url{http://www.alicebot.org/TR/2005/WD-aiml/}, 2005.
+
+\bibitem{turtlescript}
+C.~S. Ananian.
+\newblock \url{http://cscott.net/Projects/TurtleScript/}, 2011.
+
+\bibitem{bizzocchi:narrative}
+J.~Bizzocchi.
+\newblock Games and narrative: An analytical framework.
+\newblock {\em Loading---the Journal of the Canadian Games Studies
+  Association}, 1(1), 2007.
+
+\bibitem{boyd:stories}
+B.~Boyd.
+\newblock {\em On the Origin of Stories: Evolution, Cognition, and Fiction}.
+\newblock Belknap Press of Harvard University Press, 2009.
+
+\bibitem{chang:tinkrbook}
+A.~Chang and C.~Breazeal.
+\newblock {TinkRBook}: Shared reading interfaces for storytelling.
+\newblock In {\em Proc. of the 10th Int'l Conf. on Interaction Design and
+  Children (IDC '11)}, pages 145--148. ACM, June 2011.
+
+\bibitem{don:narrative}
+A.~Don.
+\newblock Narrative and the interface.
+\newblock In B.~Laurel, editor, {\em The Art of Human-Computer Interface
+  Design}, pages 383--391. Addison-Wesley, Reading, MA, 1990.
+
+\bibitem{strips}
+R.~E. Fikes and N.~J. Nilsson.
+\newblock {STRIPS}: A new approach to the application of theorem proving to
+  problem solving.
+\newblock {\em Artificial Intelligence}, 2(3--4):189--208, 1971.
+
+\bibitem{gardner:mi}
+H.~E. Gardner.
+\newblock {\em Frames Of Mind: The Theory of Multiple Intelligences}.
+\newblock Basic Books, Dec. 1983.
+
+\bibitem{lebowitz:universe85}
+M.~Lebowitz.
+\newblock Story-telling as planning and learning.
+\newblock {\em Poetics}, 14(6):483--502, Dec. 1985.
+
+\bibitem{montfort:curveship}
+N.~Montfort.
+\newblock Curveship's automatic narrative variation.
+\newblock In {\em Proc. of the 6th Int'l Conf. on the Foundations of Digital
+  Games (FDG '11)}, pages 211--218, June 2011.
+
+\bibitem{powers:appleguide}
+J.~Powers.
+\newblock Giving users help with {A}pple {G}uide.
+\newblock {\em develop}, 18, June 1994.
+\newblock Apple Computer, Inc.
+
+\bibitem{riedl:2008}
+M.~O. Riedl, A.~Stern, D.~Dini, and J.~Alderman.
+\newblock Dynamic experience management in virtual worlds for entertainment,
+  education, and training.
+\newblock {\em Int'l Trans. on Systems Science and Applications}, 3(1), 2008.
+
+\bibitem{riedl:planning}
+M.~O. Riedl and R.~M. Young.
+\newblock Narrative planning: Balancing plot and character.
+\newblock {\em Journal of Artificial Intelligence Research}, 39:217--267, 2010.
+
+\bibitem{quasiquote}
+M.~Samuel.
+\newblock Ecmascript quasi-literals.
+\newblock \url{http://wiki.ecmascript.org/doku.php?id=harmony:quasis}, 2012.
+
+\bibitem{skorupski:2009}
+J.~Skorupski and M.~Mateas.
+\newblock Interactive story generation for writers: Lessons learned from the
+  {W}ide {R}uled authoring tool.
+\newblock In {\em Proc. of the 8th Digital Art and Culture Conf. (DAC)},
+  Irvine, CA, Dec. 2009.
+
+\bibitem{stefanakis:portfolios}
+E.~Stefanakis.
+\newblock {\em Multiple Intelligences and Portfolios: A Window Into The
+  Learner's Mind}.
+\newblock Heinemann, 2002.
+
+\bibitem{sugar}
+{Sugar Labs}.
+\newblock \url{http://sugarlabs.org}.
+
+\bibitem{turkle:alone}
+S.~Turkle.
+\newblock {\em Alone Together: Why We Expect More From Technology and Less From
+  Each Other}.
+\newblock Basic Books, 2011.
+
+\bibitem{wilcox:2010}
+B.~Wilcox.
+\newblock Beyond {F}a\c{c}ade: Pattern matching for natural language
+  applications.
+\newblock
+  \url{http://chatscript.sourceforge.net/Documentation/Pattern_Matching_for_Natural_Language_Applications.pdf},
+  Feb. 2011.
+
+\end{thebibliography}
+
 %
-% ACM needs 'a single self-contained file'!
-%
-\balancecolumns
+%\balancecolumns
 % That's all folks!

### 2012-04-24 paper.tex
Fix an overfull hbox in the categories list.
Use underline for emphasis in italicized text, instead of lack-of-italics.
Rewrite figure 2 caption to address reviewer's comments (ie, define "JSON").

@@
 % A category with the (minimum) three required fields
-\category{K.3.1}{Computers and Education}{Computer
-  Uses in Education}[computer-managed instruction]
 %\category{I.2.7}{Artificial Intelligence}{Natural Language
@@
   Interfaces}[narrative interfaces]
+\category{K.3.1}{Computers and Education}{Computer
+  Uses in Education}[computer-man\-aged instruction]
 \terms{Design, Human Factors}
@@
 \emph{``Good work!  I see a little red \underline{here}.  Can you find
-something \emph{big} and red?''}
+something \underline{big} and red?''}
 % avoid excessive/unmerited praise
@@
 \end{verbatim}
-\caption{Excerpt of story module.  JSON syntax has been extended with
+\caption{Story module excerpt written in JSON
+  (\texttt{http://json.org}) extended with
  quasiquote~\cite{quasiquote}.}\label{fig:rules}

### 2012-04-24 paper.tex
"beginning" -> "starting" to fix an overfull hbox.

@@
 We have described the design of a novel narrative direct-interface
-system for education, beginning with literacy, which is personalized for and grows with
+system for education, starting with literacy, which is personalized for and grows with
 its child owner.  Personalized direct interface engages the child, and
